---
title: "ctdnaTM — Full Function Tour"
subtitle: "Every function, every key parameter · v0.19.0"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    code_folding: show
    fig_width: 7
    fig_height: 4.5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{r libs}
library(ctdnaTM)
library(ggplot2)
```

> **About this document.** This is the function-by-function tour of
> `ctdnaTM` v0.19.0. Every exported function is called at least once,
> with the main parameters demonstrated and a short explanation of
> what each does, what the parameters mean, and (where relevant) what
> the statistics mean. Use the TOC to jump anywhere.
>

---

# 1. Discovery: `ctdna_help()`

When you can't remember which function does what, start here.

`ctdna_help(category, search, show_args, width)`

- `category`: restrict output to one group (partial match: `"plot"`, `"stat"`, `"helper"`, ...)
- `search`: regex applied to both function names and summaries
- `show_args`: TRUE prints argument names; FALSE gives a compact view
- `width`: line width for wrapping (default 78)

Returns invisibly a data frame so you can use it programmatically.

```{r help-compact}
# Compact full listing, one line per function
df <- ctdna_help(show_args = FALSE, width = 90)
```

```{r help-categories, eval = FALSE}
ctdna_help("Helpers")              # everything in the Helpers group
ctdna_help("plot")                 # all plot functions (partial match)
ctdna_help(search = "RECIST")      # functions mentioning RECIST anywhere
```

```{r help-as-df}
df <- ctdna_help(show_args = FALSE)
table(df$category)
```

---

# 2. Configuration

The package has a single configuration registry; tune it once at the
top of your script.

## 2.1 `ctdna_opts(..., load, .reset)`

Five usage modes:

```{r opts-print, eval = FALSE}
ctdna_opts()                       # print full registry
```

```{r opts-get}
ctdna_opts("tf")                                   # get one
ctdna_opts(c("tf","maxaf","display_floor"))        # get several
```

```{r opts-set}
ctdna_opts(tf = "TF_v3", display_floor = 5e-4)     # set values
ctdna_opts(c("tf","display_floor"))                # verify
ctdna_opts(.reset = TRUE)                          # reset to defaults
```

```{r opts-load}
# Snapshot, mutate, restore — the `load =` form is the simple version
ctdna_opts(tf = "TF_methyl_v3", maxaf = "MAX_AF")
cfg <- ctdna_opts()
ctdna_opts(.reset = TRUE)
ctdna_opts(load = cfg)
ctdna_opts("tf")
ctdna_opts(.reset = TRUE)
```

**Key registry groups.** Column names (`subject`, `time`, `baseline`,
`dose`, `recist`, `tf`, `maxaf`, ...); numeric thresholds
(`display_floor` = 0.001, `loq` = 0.01, `alpha` = 0.05); RECIST value
lists (`recist_crpr`, `recist_ucrpr`, ...); statistics defaults
(`pairwise_test`, `cor_method`); plot defaults (`stat_position`,
`legend_position`, `show_stats`, `show_n`); multi-modal column names
(`gene`, `expression`, `tmb`, ...); `ctdna_metric` default
(`"best_ratio"`).

## 2.2 `ctdna_platforms()`

Lists supported vendor/technology combinations that `ctdna_prepare()`
understands.

```{r platforms}
ctdna_platforms()
```

---

# 3. Mock data

These two generators produce realistic test data. Both accept `seed`
for reproducibility.

## 3.1 `ctdna_make_mock_study(n_patients, visits, cancer_types, doses, n_rnaseq_genes, study_id, seed)`

Simulates a complete Guardant Health Infinity vendor bundle — five
vendor files plus clinical, IHC, WES, and RNA-seq. Built-in biology:
RECIST correlates with TF reduction, dose-response is real, mutational
landscape is plausible.

```{r mock-study}
sim <- ctdna_make_mock_study(
  n_patients     = 30,
  visits         = c("C1D1","C2D1","C3D1","C5D1","EOT"),
  cancer_types   = c("NSCLC","CRC","BRCA","mCRPC","GBM"),
  doses          = c("Low","Mid","High"),
  n_rnaseq_genes = 800,
  study_id       = "STUDY01",
  seed           = 42
)
names(sim)
```

## 3.2 `ctdna_make_example_multimodal(n_subjects, seed, vendor, technology)`

Faster, canonical-shape mock frames — best for prototyping plots.

```{r mock-mm}
mm <- ctdna_make_example_multimodal(n_subjects = 30, seed = 42)
names(mm)
attr(mm$expression, "units")        # tags the units so to_log2tpm() can canonicalize
```

```{r alias-ctdna}
ctdna_df <- mm$ctdna  # used in the plot demos below
```

---

# 4. Vendor → canonical

## 4.1 `ctdna_prepare(infinity_report, tf_change, panel_74_response, panel_500_response, methylation_response, clinical, vendor, technology, qc_filter, verbose)`

The flagship transform. Takes raw Infinity files + a clinical table
and emits canonical long-format frames every plot consumes.

```{r prepare}
prepped <- ctdna_prepare(
  infinity_report    = sim$infinity_report,
  tf_change          = sim$tf_change,
  panel_74_response  = sim$panel_74_response,
  panel_500_response = sim$panel_500_response,
  clinical           = sim$clinical,
  qc_filter          = TRUE,
  verbose            = FALSE
)
names(prepped)
head(prepped$ctdna, 2)
```

> Prepared data uses vendor visit names (`C1D1`, `C2D1`, ...). The
> default baseline label in the package is `"Baseline"`, matching
> `mm$ctdna`. For real prepared data, tell the package which visit is
> baseline once with `ctdna_opts(baseline = "C1D1")`.

## 4.2 `ctdna_assemble_modalities(ctdna, genomic_74, genomic_500, expression, mutations, tmb, ihc, strict)`

Validates required columns across modalities and assembles a list
with canonical schemas. Auto-converts expression to log2(TPM+1).

```{r assemble}
db <- ctdna_assemble_modalities(
  ctdna       = mm$ctdna,
  genomic_74  = mm$genomic_74,
  genomic_500 = mm$genomic_500,
  expression  = mm$expression,
  tmb         = mm$tmb,
  ihc         = mm$ihc,
  strict      = TRUE
)
names(db)
```

## 4.3 `ctdna_to_log2tpm(df)`

Canonical unit throughout the package is **log2(TPM+1)**. The helper
detects raw TPM via `attr(df, "units")` (set explicitly) or by value
range (max > 30 → assumed TPM).

```{r to-log2tpm}
expr <- data.frame(
  subject_id = c("S1","S2","S3"),
  gene       = "PDL1",
  expression = c(0, 5.4, 412)
)
attr(expr, "units") <- "tpm"
out <- ctdna_to_log2tpm(expr)
out$expression                         # log2(1)=0, log2(6.4)~2.68, log2(413)~8.69
attr(out, "units")
```

---

# 5. Helpers

## 5.1 `ctdna_floor_tf(x, at)`

Replace zeros and below-LOQ values with a display floor so they render
on log plots.

```{r floor-tf}
ctdna_floor_tf(c(0, 1e-5, 0.001, 0.5))             # default floor
ctdna_floor_tf(c(0, 1e-5, 0.001, 0.5), at = 1e-3)  # custom floor
```

## 5.2 `ctdna_stratify_recist(x, scheme)`

Collapse RECIST values into clinically useful strata. **The chosen
`scheme` drives both colours/groups in plots AND statistical
comparisons.**

- `"two"` — R (CR/PR/uCR/uPR) vs NR (SD/PD/NE/NA)
- `"three"` — CR/PR vs SD vs PD/NE/NA  *(uCR/PR considered with CR/PR by default if absent)*
- `"four"` — CR/PR vs uCR/PR vs SD vs PD/NE/NA

```{r stratify}
r <- c("CR","PR","uCR","uPR","SD","PD","NE","NA")
ctdna_stratify_recist(r, scheme = "two")
ctdna_stratify_recist(r, scheme = "three")
ctdna_stratify_recist(r, scheme = "four")
```

## 5.3 `ctdna_ratio(df)`

Adds per-subject `baseline_tf` and per-visit `ratio = TF_visit / TF_baseline`.

```{r ratio}
head(ctdna_ratio(ctdna_df)[, c("subject_id","time_point","methylTF",
                                "baseline_tf","ratio")], 4)
```

## 5.4 `ctdna_summary(df)`

Collapse a longitudinal frame to one row per subject (baseline TF, best
on-treatment ratio, last TF, metadata).

```{r summary}
head(ctdna_summary(ctdna_df), 3)
```

## 5.5 `ctdna_metric_at(df, metric, at)`

Single source of truth for per-subject ctDNA summary used in
cross-modality plots and D3.

**`metric`:**

- `"ratio"` — `TF_visit / TF_baseline`. `0.1` means TF dropped to 10% of baseline.
- `"tf"` — raw TF at the visit.
- `"pct_change"` — `100 · (TF_visit − TF_baseline) / TF_baseline`. Negative = reduction.

**`at`:**

- `"best"` — per-subject deepest response (lowest ratio).
- `"last"` — chronologically last on-treatment visit per subject.
- a visit name (e.g. `"C5D1"`) — that exact visit.

```{r metric-at}
head(ctdna_metric_at(ctdna_df), 3)
head(ctdna_metric_at(ctdna_df, metric = "tf",         at = "C5D1"), 3)
head(ctdna_metric_at(ctdna_df, metric = "pct_change", at = "last"), 3)
```

## 5.6 `ctdna_compute_mr(df, at, threshold)`

Per-subject Molecular Response classifier: ratio at `at` is compared
against `threshold` (default 0.5 = "TF dropped by at least 50%").
Returns one row per subject with the `MR` column added (`"Response"`
or `"NonResponse"`).

```{r compute-mr}
head(ctdna_compute_mr(ctdna_df), 3)
head(ctdna_compute_mr(ctdna_df, at = "C5D1"), 3)
head(ctdna_compute_mr(ctdna_df, threshold = 0.3), 3)  # stricter — 70% drop required
```

## 5.7 `ctdna_merge_modalities(ctdna_df, other_df, by)`

Per-subject join of ctDNA summary with another modality.

```{r merge}
head(ctdna_merge_modalities(ctdna_df, mm$tmb, by = "subject_id"), 3)
```

## 5.8 `ctdna_filter_alterations(df, ...)`

Unified alteration filter. Every parameter accepts `FALSE` to disable
that filter. Works on both vendor (Guardant Infinity) and canonical
schemas.

```{r filter-defaults}
cat("Total alterations:", nrow(sim$infinity_report), "\n")
filtered <- ctdna_filter_alterations(
  sim$infinity_report,
  variant_type      = c("SNV","Indel","CNV","Fusion","LGR"),
  somatic_status    = c("Somatic","Likely somatic"),
  functional_impact = c("High","Moderate"),
  allelic_status    = FALSE,        # disable this filter entirely
  max_gnomad_af     = 0.01,         # gnomAD population frequency ceiling
  max_kg_af         = 0.01,
  min_vaf           = 0.5,          # tumor-fraction ceiling for somatic calls
  min_cn_amp        = 4,            # amplification threshold (copy number)
  max_cn_loss       = 1,            # loss threshold
  keep_wt           = FALSE,
  verbose           = FALSE
)
cat("After standard filter:", nrow(filtered), "\n")
```

```{r filter-clinvar}
# Pathogenic-only filter (turn off functional_impact so ClinVar takes precedence)
ctdna_filter_alterations(
  sim$infinity_report,
  clinvar           = c("Pathogenic","Likely pathogenic"),
  functional_impact = FALSE,
  verbose           = FALSE
) |> nrow()
```

```{r filter-panel}
# Restrict to one panel (e.g. just the 74-gene calls)
ctdna_filter_alterations(
  sim$infinity_report,
  panel  = "74genes",
  verbose = FALSE
) |> nrow()
```

## 5.9 `ctdna_filter_opts(...)` — column-by-column filter registry

The registry is a column-by-column inventory of every filterable
Guardant Infinity column. Each entry is
`list(enabled = TRUE/FALSE, op = "%in%"/">="/"<="/"==", value = ..., desc = "...")`.
Toggle columns on and off, change thresholds, save and restore
snapshots — same five-mode pattern as `ctdna_opts()`.

### Inspecting the registry

```{r filter-opts-print, eval = FALSE}
ctdna_filter_opts()                            # print all 28 entries
ctdna_filter_opts("VAF_percentage")            # get one entry
ctdna_filter_opts(c("VAF_percentage","ClinVar"))  # get several
```

### Modifying entries

You can pass a partial list — fields you don't mention keep their
current value.

```{r filter-opts-set}
ctdna_filter_opts(.reset = TRUE)               # start clean

# Turn on VAF threshold at 1.5%; tighten Functional_impact to High only;
# add a Mol_count floor.
ctdna_filter_opts(
  VAF_percentage    = list(enabled = TRUE, value = 1.5),
  Functional_impact = list(value = "High"),
  Mol_count         = list(enabled = TRUE, value = 100))

ctdna_filter_opts("VAF_percentage")
```

### Save & restore snapshots

```{r filter-opts-snapshot}
cfg <- ctdna_filter_opts()
ctdna_filter_opts(.reset = TRUE)
ctdna_filter_opts(load = cfg)
ctdna_filter_opts("VAF_percentage")$enabled
```

```{r filter-opts-finalreset, include = FALSE}
ctdna_filter_opts(.reset = TRUE)
```

### What columns are filterable?

Categorical (`%in%` operator, allowlist): `Sample_status`, `Variant_type`,
`Functional_impact`, `Molecular_consequence`, `Somatic_status`,
`ClinVar`, `Mutant_allele_status`, `TMB_category`, `Cancertype`,
`Visit_name`, `Patient_ID`, `Gene`, `Chromosome`, `MSI_High`,
`ctDNA_detection_status`, `Splice_effect`, `Alleletype`, `COSMIC`,
`dbSNP`.

Numeric (`>=` / `<=` / `==`, threshold): `VAF_percentage`,
`Genomic_max_VAF_percentage`, `Mol_count`, `Copy_number`, `TMB_score`,
`HRD_score`, `Methylation_tumor_fraction_percentage`, `cfDNA_ng`,
`Plasma_ml_input`.

## 5.11 Three-level filter system

The filter registry from §5.9 is just one half of the story. The
package gives you **three levels** of filtering to choose from when
calling any function that touches alterations.

### The three levels

**Level 1 — default general.** Shipped with the package. Sensible
defaults applied uniformly to every gene. Always preserved as a hard
reset target so you can undo any modification.

**Level 2 — modified general.** You edit the general filter (the
§5.9 registry) and these become the active rules applied uniformly to
every gene by every function.

**Level 3 — named specific schemes.** You create one or more named
schemes (e.g. `schemeCAD`, `schemeStrict`). Each scheme attaches
per-gene-set OVERRIDES on top of the general filter. Genes not in any
of the scheme's gene sets fall back to the general filter for that
scheme.

### At call-time

Every function that uses alterations takes a single argument:

- `filter_scheme = NULL` (default) → use the current general filter
  (level 1 or 2, whichever state it is in)
- `filter_scheme = "name"` → use that named specific scheme (level 3)

### Dynamic vs frozen schemes

When you create a scheme you choose its **mode**:

- **dynamic** (default) — the scheme stores only the per-gene-set
  overrides. When applied, the active general filter is read at that
  moment. If you later edit general, dynamic schemes automatically
  pick up the change for gene sets they do not override.
- **frozen** (`frozen = TRUE` at create time, or
  `ctdna_filter_scheme_freeze(name)` later) — the scheme snapshots the
  general filter NOW and locks it in. Use this when you need exact
  reproducibility against future edits.

### Reset back to defaults

`ctdna_filter_reset_general()` restores the active general to factory
defaults. Specific schemes are untouched (dynamic ones re-resolve on
next use; frozen ones are independent by design).

### Worked example

```{r filter-scheme-demo}
# Start from a known state
ctdna_filter_reset_general()
n_default <- nrow(ctdna_filter_apply_scheme(sim$infinity_report))
cat("Level 1 - default general:", n_default, "rows\n")

# Level 2: tighten general
ctdna_filter_opts(VAF_percentage = list(enabled = TRUE, value = 2.0))
n_mod <- nrow(ctdna_filter_apply_scheme(sim$infinity_report))
cat("Level 2 - modified general (VAF >= 2):", n_mod, "rows\n")

# Level 3: named specific scheme with different rules per gene set
ctdna_filter_scheme_create(
  "schemeCAD",
  gene_sets = list(
    HRR     = c("BRCA1","BRCA2","ATM","PALB2"),
    TP53    = "TP53",
    RTK_RAS = c("KRAS","EGFR","BRAF","MET")),
  overrides = list(
    HRR     = list(ClinVar = list(enabled = TRUE,
                                    value = c("Pathogenic","Likely pathogenic"))),
    TP53    = list(Mutant_allele_status = list(enabled = TRUE,
                                                  value = "Homozygous")),
    RTK_RAS = list(VAF_percentage = list(enabled = TRUE, value = 5))))
n_spec <- nrow(ctdna_filter_apply_scheme(sim$infinity_report,
                                           filter_scheme = "schemeCAD"))
cat("Level 3 - schemeCAD:", n_spec, "rows\n")

# List schemes
ctdna_filter_scheme_list()
```

### Dynamic vs frozen demonstration

```{r filter-scheme-frozen}
# Two schemes with identical contents - one dynamic, one frozen
ctdna_filter_scheme_create("dyn_demo",
  gene_sets = list(HRR = c("BRCA1","BRCA2")),
  overrides = list())
ctdna_filter_scheme_create("frz_demo",
  gene_sets = list(HRR = c("BRCA1","BRCA2")),
  overrides = list(),
  frozen = TRUE)

# Now tighten general AFTER both were created
ctdna_filter_opts(VAF_percentage = list(enabled = TRUE, value = 10))

n_dyn <- nrow(ctdna_filter_apply_scheme(sim$infinity_report,
                                          filter_scheme = "dyn_demo"))
n_frz <- nrow(ctdna_filter_apply_scheme(sim$infinity_report,
                                          filter_scheme = "frz_demo"))
cat("dyn_demo (re-reads general):", n_dyn, "rows\n")
cat("frz_demo (snapshot at create time):", n_frz, "rows\n")

# Clean up for the rest of the tour
ctdna_filter_reset_general()
ctdna_filter_scheme_delete("schemeCAD")
ctdna_filter_scheme_delete("dyn_demo")
ctdna_filter_scheme_delete("frz_demo")
```

### The `has_value` operator

For columns like `dbSNP` and `COSMIC` you often care only about
whether a value exists, not what it is. The `has_value` operator
keeps rows with any value (`value = TRUE`) or rows that are missing
(`value = FALSE`):

```{r filter-has-value}
# Keep only NOVEL variants (no dbSNP annotation)
ctdna_filter_opts(dbSNP = list(enabled = TRUE, op = "has_value",
                                 value = FALSE))
n_novel <- nrow(ctdna_filter_apply_scheme(sim$infinity_report))
cat("Novel (no dbSNP):", n_novel, "rows\n")

# Keep only ANNOTATED variants (has any dbSNP value)
ctdna_filter_opts(dbSNP = list(enabled = TRUE, op = "has_value",
                                 value = TRUE))
n_anno <- nrow(ctdna_filter_apply_scheme(sim$infinity_report))
cat("Annotated (has dbSNP):", n_anno, "rows\n")

ctdna_filter_reset_general()
```

### Lifecycle helpers

| Function | Purpose |
|---|---|
| `ctdna_filter_reset_general()` | Restore active general to factory defaults |
| `ctdna_filter_scheme_create(name, gene_sets, overrides, frozen)` | Define a specific scheme |
| `ctdna_filter_scheme_list()` | List all defined schemes |
| `ctdna_filter_scheme_get(name)` | Retrieve a scheme object |
| `ctdna_filter_scheme_delete(name)` | Remove a scheme |
| `ctdna_filter_scheme_freeze(name)` | Convert dynamic → frozen |
| `ctdna_filter_scheme_save(file)` | Save schemes + general to RDS |
| `ctdna_filter_scheme_load(file)` | Restore from RDS |

### Where `filter_scheme` is accepted

`ctdna_plot_oncoprint`, `ctdna_plot_alteration_grid`,
`ctdna_plot_vs_mutation`, `ctdna_mutation_concordance`.

---

# 6. Statistics

## 6.1 `ctdna_pval_label(p)`

Vectorized p-value formatter: `"p<0.001"`, `"p=0.012"`, etc.

```{r pval}
ctdna_pval_label(c(1e-6, 0.003, 0.045, 0.5, NA))
```

## 6.2 `ctdna_run_group_test(data, value, group, method)`

Auto-selects the right test for the data. `method = "auto"` (default)
picks: Wilcoxon (2 groups) or Kruskal-Wallis (3+ groups) if
`pairwise_test = "wilcox"`; t-test or ANOVA if `pairwise_test = "t"`.
Force a specific test with `method = "wilcox" / "kruskal" / "t" / "anova"`.

**What you get back:** `name` (test used), `p` (p-value), `n`.

```{r run-group-test}
base <- subset(ctdna_df, time_point == "Baseline")
ctdna_run_group_test(base, value = "methylTF", group = "RECIST")  # 6+ levels → KW
ctdna_run_group_test(base, value = "methylTF", group = "dose",
                      method = "kruskal")
```

## 6.3 `ctdna_overall_group_test(data, value, group, method)`

Single overall test (Kruskal-Wallis or one-way ANOVA).

```{r overall}
ctdna_overall_group_test(base, value = "methylTF", group = "RECIST")
```

## 6.4 `ctdna_pairwise_ranksum(data, value, group, method)`

All-pairwise Wilcoxon (or t-test) with one row per pair.

```{r pairwise}
ctdna_pairwise_ranksum(base, value = "methylTF", group = "RECIST",
                        method = "wilcox")[1:5, ]
```

## 6.5 `ctdna_cor_test(x, y, method)`

Spearman (default), Pearson, or Kendall correlation with rho, p-value,
and n. **Spearman** is the safe default for ctDNA fractions because
they are skewed and have a hard floor at the LOQ.

```{r cor}
ctdna_cor_test(base$methylTF, base$maxAF, method = "spearman")
ctdna_cor_test(base$methylTF, base$maxAF, method = "pearson")
```

## 6.6 `ctdna_paired_test(x, y, method)`

Paired comparison (baseline vs on-treatment per subject). `method =
"wilcox"` (paired Wilcoxon signed-rank, default) or `"t"` (paired t).

```{r paired}
ctdna_paired_test(c(1,2,3,4), c(1.1, 2.3, 2.8, 4.2), method = "wilcox")
```

## 6.7 `ctdna_cohen_kappa(tab)`

Cohen's kappa for a 2×2 (or k×k) contingency — chance-adjusted
agreement. Returns kappa, the observed proportion of agreement, and
the chance-expected proportion.

```{r kappa}
tab <- matrix(c(12, 3, 1, 9), 2, 2,
              dimnames = list(MR = c("Resp","NonResp"),
                              RECIST = c("CR/PR","SD/PD")))
ctdna_cohen_kappa(tab)
```

## 6.8 `ctdna_concordance_test(tab, method)`

Agreement % + Cohen's kappa + (2×2 only) McNemar p-value in one call.

```{r concordance}
ctdna_concordance_test(tab, method = "both")
```

---

# 7. Display options every plot accepts

| Option | Default | Effect |
|---|---|---|
| `scheme` | varies | RECIST stratification: `"two"` / `"three"` / `"four"` (plus `"raw"` for oncoprint / alt grid) |
| `stat_position` | `"subtitle"` | `"subtitle"` / `"caption"` / `"on_plot"` / `"none"` |
| `legend_position` | `"right"` | `"right"` / `"left"` / `"top"` / `"bottom"` / `"none"` |
| `show_stats` | `TRUE` | Compute and display the stats label |
| `show_n` | `TRUE` | Annotate `n=N` under each x-tick |
| `title`, `subtitle`, `caption`, `xlab`, `ylab` | per-plot | Standard labels |
| `point_size`, `point_alpha` | 2, 0.85 | Point styling |

Set as global defaults with
`ctdna_opts(stat_position = "caption", legend_position = "bottom")`.

```{r stat-positions, fig.height = 3.5}
ctdna_plot_vs_tmb(ctdna_df, mm$tmb,
                    stat_position = "subtitle",
                    title = "stat_position = 'subtitle' (default)")
ctdna_plot_vs_tmb(ctdna_df, mm$tmb,
                    stat_position = "caption",
                    title = "stat_position = 'caption'")
ctdna_plot_vs_tmb(ctdna_df, mm$tmb,
                    stat_position = "on_plot",
                    title = "stat_position = 'on_plot'")
```

```{r legend-position}
ctdna_plot_baseline(ctdna_df, group = "RECIST", vars = "methylTF",
                     legend_position = "bottom",
                     title = "legend_position = 'bottom'")
```

---

# 8. The 8 core deliverables (D1–D8)

## D1 — `ctdna_plot_baseline(df, group, vars, recist_shape, scheme, stat, ...)`

Boxplot of one or more baseline variables grouped by any clinical
factor. When `group = "RECIST"`, the RECIST palette is used
automatically; `scheme` controls both the collapse and the statistics.

```{r D1-recist}
ctdna_plot_baseline(ctdna_df,
                     group  = "RECIST",
                     vars   = "methylTF",
                     scheme = "three",
                     stat   = "auto",
                     title  = "D1 — Baseline methylTF by RECIST (scheme=three)")
```

```{r D1-multi-vars}
# Multiple variables in side-by-side facets, grouped by dose
ctdna_plot_baseline(ctdna_df,
                     group = "dose",
                     vars  = c("methylTF","maxAF"),
                     stat  = "kruskal",
                     title = "D1 — Two variables faceted, grouped by dose")
```

## D2 — `ctdna_plot_baseline_dose_recist(df, scheme, stat, ...)`

Baseline methylTF dodged by dose × RECIST. Pairwise comparisons within
each dose, scheme-aware.

```{r D2}
ctdna_plot_baseline_dose_recist(ctdna_df,
                                  scheme = "three",
                                  stat   = "wilcox",
                                  title  = "D2 — Baseline by dose × RECIST")
```

## D3 — `ctdna_plot_reduction(df, scheme, at, metric, mr_threshold, stat, ...)`

Per-patient ctDNA reduction; the most flexible deliverable.

- `metric`: see §5.5 (`ratio` / `tf` / `pct_change`)
- `at`: see §5.5 (`best` / `last` / named visit)
- `mr_threshold`: ratio cutoff; a dashed horizontal line is drawn when `metric = "ratio"`. Patients below the line are MR Responders.

```{r D3-default}
ctdna_plot_reduction(ctdna_df,
                       scheme       = "three",
                       at           = "best",
                       metric       = "ratio",
                       mr_threshold = 0.5,
                       stat         = "wilcox",
                       title = "D3 — ratio @ best, MR threshold = 0.5")
```

```{r D3-variations, fig.height = 4}
ctdna_plot_reduction(ctdna_df, scheme = "three",
  metric = "ratio", at = "C5D1",
  title = "D3 — ratio at a specific visit (C5D1)")

ctdna_plot_reduction(ctdna_df, scheme = "three",
  metric = "tf", at = "last",
  title = "D3 — absolute TF at the last on-treatment visit")

ctdna_plot_reduction(ctdna_df, scheme = "three",
  metric = "pct_change", at = "best",
  title = "D3 — percent change at best (negative = reduction)")
```

## D4 — `ctdna_plot_mr_recist(df, scheme, mr_timepoint, mr_threshold, stat, ...)`

MR ↔ RECIST 2×2 concordance heatmap + publication-style contingency
table.

- `mr_timepoint`: `"best"`, a named visit (`"C5D1"`), or `"use_column"`
  (uses a pre-computed `MR` column).
- `mr_threshold`: ratio cutoff for the MR call.
- `stat`: `"both"` (agreement + kappa, default), `"kappa"`,
  `"agreement"`, `"mcnemar"`, `"none"`.

```{r D4-best}
mr <- ctdna_plot_mr_recist(ctdna_df,
                             mr_timepoint = "best",
                             mr_threshold = 0.5,
                             stat = "both",
                             title = "D4 — MR / RECIST (MR @ best)")
mr$plot
mr$table_plot
mr$agreement; mr$kappa; mr$mcnemar_p
```

```{r D4-stricter}
ctdna_plot_mr_recist(ctdna_df,
                      mr_timepoint = "best",
                      mr_threshold = 0.3,
                      title = "D4 — Stricter threshold (90% drop required)")$plot
```

## D5 — `ctdna_plot_longitudinal(df, stat, recist_color, scheme, ...)`

Per-subject spaghetti trajectory + thick median per dose. `stat`
default is `"paired_wilcox"` (baseline vs each on-treatment visit per
subject). Optional `recist_color = TRUE` colours subject lines by
RECIST stratum.

```{r D5-default}
ctdna_plot_longitudinal(ctdna_df,
                          stat  = "paired_wilcox",
                          title = "D5 — Longitudinal trajectory by dose")
```

```{r D5-recist}
ctdna_plot_longitudinal(ctdna_df,
                          recist_color = TRUE, scheme = "three",
                          title = "D5 — Subject lines coloured by RECIST")
```

## D6 — `ctdna_plot_longitudinal_recist(df, scheme, stat, ...)`

Same longitudinal view but faceted by dose × RECIST. Per-stratum
Kruskal-Wallis (abbreviated `KW` in the subtitle).

```{r D6}
ctdna_plot_longitudinal_recist(ctdna_df,
                                 scheme = "three",
                                 stat   = "wilcox",
                                 title  = "D6 — Longitudinal by dose × RECIST")
```

## D7 — `ctdna_plot_reduction_vs_tumor(df, scheme, stat, ...)`

Scatter of per-subject best ctDNA-reduction ratio vs % change in
target lesions. **Visual encoding: RECIST = colour, dose = shape.**
Dashed red line at the RECIST PR threshold (−30% target lesions).

```{r D7}
ctdna_plot_reduction_vs_tumor(ctdna_df,
                                scheme = "three",
                                stat   = "spearman",
                                title  = "D7 — Reduction vs % tumor change")
```

## D8 — `ctdna_plot_mut_methyl(df, stat, recist_shape, scheme, ...)`

Per-baseline maxAF vs methylTF concordance. Discordant points (log10
distance > `discord_log10`, default 1) are highlighted red. Optional
`recist_shape = TRUE` overlays RECIST as point shape.

```{r D8-default}
ctdna_plot_mut_methyl(ctdna_df,
                        stat  = "spearman",
                        title = "D8 — maxAF vs methylTF")
```

```{r D8-recist}
ctdna_plot_mut_methyl(ctdna_df,
                        recist_shape = TRUE, scheme = "three",
                        title = "D8 — with RECIST as point shape")
```

---

# 9. Cross-modality

All four cross-modal plots take `ctdna_df`, the modality frame, and a
ctDNA `metric` (default `best_ratio` — the lowest TF ratio per subject
across on-treatment visits, i.e. their deepest response).

## 9.1 `ctdna_plot_vs_expression(ctdna_df, expr_df, gene, metric, log_x, recist_color, scheme, stat, ...)`

- `gene`: which gene's expression to compare against.
- `log_x`: log-scale the ctDNA axis (default TRUE).
- `recist_color`: colour points by RECIST stratum.
- `stat`: `"spearman"` (recommended), `"pearson"`, `"kendall"`, `"none"`.

```{r vs-expr}
ctdna_plot_vs_expression(ctdna_df, mm$expression,
                          gene         = "PDL1",
                          stat         = "spearman",
                          recist_color = TRUE,
                          scheme       = "two",
                          title        = "ctDNA best_ratio vs PDL1 expression")
```

## 9.2 `ctdna_plot_vs_ihc(ctdna_df, ihc_df, marker, metric, recist_color, scheme, stat, ...)`

```{r vs-ihc}
ctdna_plot_vs_ihc(ctdna_df, mm$ihc,
                    marker       = "PDL1",
                    stat         = "spearman",
                    recist_color = TRUE,
                    scheme       = "three",
                    title        = "ctDNA best_ratio vs PDL1 IHC")
```

## 9.3 `ctdna_plot_vs_mutation(ctdna_df, mut_df, gene, panel, by, metric, recist_facet, scheme, stat, ...)`

- `by`: `"mutation_status"` (mut vs wt) or `"alteration_type"`.
- `recist_facet`: facet the boxplot by RECIST stratum.

```{r vs-mut}
ctdna_plot_vs_mutation(ctdna_df, mm$genomic_74,
                         gene  = "TP53",
                         panel = "74genes",
                         stat  = "wilcox",
                         title = "ctDNA best_ratio by TP53 status")
```

```{r vs-mut-facet}
ctdna_plot_vs_mutation(ctdna_df, mm$genomic_74,
                         gene         = "TP53",
                         panel        = "74genes",
                         recist_facet = TRUE,
                         scheme       = "two",
                         title        = "ctDNA vs TP53, faceted by RECIST stratum")
```

## 9.4 `ctdna_plot_vs_tmb(ctdna_df, tmb_df, panel, metric, tmb_col, recist_color, scheme, stat, ...)`

```{r vs-tmb}
ctdna_plot_vs_tmb(ctdna_df, mm$tmb,
                    recist_color = TRUE,
                    scheme       = "three",
                    stat         = "spearman",
                    title        = "ctDNA best_ratio vs TMB (coloured by RECIST)")
```

## 9.5 `ctdna_mutation_concordance(mut_df, panel_a, panel_b, genes, stat, ...)`

Compares per-(subject × gene) mutation calls between two panels on
the SAME patients. Useful for 74-gene vs 500-gene comparisons, or
plasma vs tissue.

**Plot:** per-gene agreement % (bars sorted ascending, `n` under each
gene name). **Stats:** overall agreement, Cohen's kappa, (2×2 only)
McNemar p.

```{r mut-conc, fig.height = 4.5}
mut_long <- rbind(mm$genomic_74, mm$genomic_500)
mc <- ctdna_mutation_concordance(
  mut_long,
  panel_a = "74genes", panel_b = "500genes",
  stat    = "both",
  title   = "74-gene vs 500-gene mutation concordance"
)
mc$plot
mc$table_plot
mc$agreement; mc$kappa
head(mc$per_gene, 3)
```

---

# 10. Genomic landscape

## 10.1 `ctdna_plot_oncoprint(df, gene_sets, patient_data, top_annotations, cancer_type, cancer_type_col, scheme, sort_genes, show_freq_bar, filter, ..., engine)`

When `ComplexHeatmap` is installed (recommended; install once via
`BiocManager::install("ComplexHeatmap")`), uses `oncoPrint()` for a
publication-quality figure with:

- right-margin per-gene barplot
- top per-patient burden bar
- stacked top-annotation tracks
- gene-set row splits (when `sort_genes = "within_set"`) or coloured left-annotation columns (otherwise)
- merged legend at the bottom

Fallback ggplot tile-plot when ComplexHeatmap isn't installed.

### Key parameters

- **`gene_sets`** — NAMED list of gene-symbol vectors. `NULL` = all
  genes; with names, each set becomes either a coloured column in the
  left annotation (global sort) or a separate horizontal band
  (within-set sort).
- **`patient_data`** — per-patient frame; required for
  `top_annotations` and `cancer_type`.
- **`top_annotations`** — character vector of `patient_data` columns
  to display as stacked tracks above the oncoprint.
- **`cancer_type`** — character vector of indication values
  (e.g. `"NSCLC"` or `c("NSCLC","HNSCC")`). Restricts the cohort to
  patients whose `cancer_type_col` value matches; useful for
  per-indication oncoprints in multi-indication studies. The title
  auto-populates to "Oncoprint - <indication>" if left at the default.
- **`cancer_type_col`** — name of the indication column in
  `patient_data`. Default `"Cancertype"`.
- **`scheme`** — `"raw"` (default) or `"two"` / `"three"` / `"four"`
  to collapse the RECIST annotation track.
- **`sort_genes`** — controls row ordering. Three modes:
    - **`"global"`** (default): rank genes by overall alteration
      frequency, ignoring set membership. Best when sets overlap or
      you want a single ranked list. Set membership shown as coloured
      left-annotation columns (a gene can belong to multiple sets).
    - **`"within_set"`**: rank genes by frequency WITHIN each set,
      with a horizontal gap separating sets. Best when sets are
      disjoint pathways you want to compare side-by-side.
    - **`"none"`**: preserve the order the user gave in `gene_sets`.
    - Legacy: `TRUE` → `"global"`, `FALSE` → `"none"`.
- **`filter`** — list of args forwarded to
  `ctdna_filter_alterations()` (variant type, ClinVar, gnomAD, ...).
- **`engine`** — selects the rendering backend.

### What is `engine`?

`ctdna_plot_oncoprint()` can render the figure two ways:

| Value | Behaviour |
|---|---|
| `"auto"` (default) | Use `ComplexHeatmap::oncoPrint()` if the package is installed; otherwise fall back to a ggplot tile plot. Recommended for almost all uses. |
| `"complexheatmap"` | Force the ComplexHeatmap engine. Errors with install instructions if missing. Pick this when you need publication-quality output and want a clear failure (not a silent downgrade) if the dependency is absent. |
| `"ggplot"` | Force the ggplot tile plot even if ComplexHeatmap is installed. Useful when you want a plain ggplot object you can compose with **patchwork** or theme further. |

### Examples

```{r oncoprint-default, fig.height = 4.5}
ctdna_plot_oncoprint(sim$infinity_report,
                      filter = list(verbose = FALSE),
                      title  = "Oncoprint - default (one ranked list)")
```

```{r oncoprint-cancer-type, fig.height = 4.5}
# Filter to one indication
ctdna_plot_oncoprint(sim$infinity_report,
                      patient_data = sim$clinical,
                      cancer_type  = "NSCLC",
                      gene_sets = list(RTK_RAS = c("KRAS","EGFR","BRAF","MET"),
                                         TP53    = "TP53"),
                      top_annotations = c("RECIST","Dose"),
                      filter = list(verbose = FALSE))
```

```{r oncoprint-cancer-multi, fig.height = 4.5}
# Multiple indications at once
ctdna_plot_oncoprint(sim$infinity_report,
                      patient_data = sim$clinical,
                      cancer_type  = c("NSCLC","CRC"),
                      filter = list(verbose = FALSE),
                      title  = "Oncoprint - NSCLC + CRC pooled")
```

```{r oncoprint-global, fig.height = 5}
# sort_genes = "global" (default): one ranked list across all sets;
# membership shown as coloured left-annotation columns.
ctdna_plot_oncoprint(sim$infinity_report,
  gene_sets = list(
    RTK_RAS = c("KRAS","NRAS","EGFR","ERBB2","BRAF","MET","RET"),
    PI3K    = c("PIK3CA","PTEN"),
    TP53    = "TP53"),
  sort_genes = "global",
  filter = list(verbose = FALSE),
  title  = "Oncoprint - sort_genes = 'global'"
)
```

```{r oncoprint-within-set, fig.height = 6}
# sort_genes = "within_set": each set is its own horizontal band with
# a visible gap; genes ranked by frequency within each band.
ctdna_plot_oncoprint(sim$infinity_report,
  gene_sets = list(
    RTK_RAS = c("KRAS","NRAS","EGFR","ERBB2","BRAF","MET","RET"),
    PI3K    = c("PIK3CA","PTEN"),
    TP53    = "TP53"),
  sort_genes = "within_set",
  filter = list(verbose = FALSE),
  title  = "Oncoprint - sort_genes = 'within_set'"
)
```

```{r oncoprint-annotated, fig.height = 6}
# Full annotated oncoprint with RECIST collapsed by scheme
ctdna_plot_oncoprint(sim$infinity_report,
  gene_sets = list(
    RTK_RAS = c("KRAS","EGFR","BRAF","MET"),
    TP53    = "TP53"
  ),
  patient_data    = sim$clinical,
  top_annotations = c("RECIST","Dose","Cancertype"),
  scheme          = "three",
  filter          = list(clinvar = c("Pathogenic","Likely pathogenic"),
                          functional_impact = FALSE,
                          verbose = FALSE),
  title           = "Oncoprint with annotations (scheme=three)"
)
```

```{r oncoprint-engine-force, eval = FALSE}
# Force the ggplot fallback (returns a plain ggplot object inside)
ctdna_plot_oncoprint(sim$infinity_report,
                      engine = "ggplot",
                      filter = list(verbose = FALSE))
```

### Filtering: general (level 1/2) and specific schemes (level 3)

The oncoprint uses the three-level filter system from §5.11. One
argument routes the cohort through any of the three levels:

- **`filter_scheme = NULL`** (default) — apply the current general
  filter to all genes. Whatever state `ctdna_filter_opts()` is in,
  whether factory default (level 1) or modified (level 2), is what
  gets applied uniformly to every gene.
- **`filter_scheme = "name"`** — apply a named specific scheme created
  with `ctdna_filter_scheme_create()`. Each gene set inside the scheme
  has its own merged filter (general + that set's overrides); genes
  not in any of the scheme's sets fall back to the general filter.

```{r oncoprint-filter-general, fig.height = 5}
# Reset to a known state
ctdna_filter_reset_general()

# Default: the active general filter applied uniformly
ctdna_plot_oncoprint(sim$infinity_report,
  gene_sets = list(
    RTK_RAS = c("KRAS","EGFR","BRAF","MET"),
    HRR     = c("BRCA1","BRCA2","ATM","PALB2"),
    TP53    = "TP53"),
  filter_scheme = NULL,                  # = general
  title = "Oncoprint - general filter only"
)
```

```{r oncoprint-filter-scheme, fig.height = 5}
# Define a specific scheme with per-gene-set overrides, then pass its
# name. RTK_RAS needs higher VAF; HRR restricted to ClinVar Path/LP;
# TP53 inherits the general filter (no override specified for it).
ctdna_filter_scheme_create("schemeStrict",
  gene_sets = list(
    RTK_RAS = c("KRAS","EGFR","BRAF","MET"),
    HRR     = c("BRCA1","BRCA2","ATM","PALB2"),
    TP53    = "TP53"),
  overrides = list(
    RTK_RAS = list(VAF_percentage = list(enabled = TRUE, value = 3.0)),
    HRR     = list(ClinVar = list(enabled = TRUE,
                                    value = c("Pathogenic","Likely pathogenic")))),
  overwrite = TRUE)

ctdna_plot_oncoprint(sim$infinity_report,
  gene_sets = list(
    RTK_RAS = c("KRAS","EGFR","BRAF","MET"),
    HRR     = c("BRCA1","BRCA2","ATM","PALB2"),
    TP53    = "TP53"),
  filter_scheme = "schemeStrict",
  sort_genes = "within_set",
  title = "Oncoprint - filter_scheme = 'schemeStrict'"
)
```

How merging works inside `schemeStrict`:

| Gene set | General filter | Override applied | Final filter for that set |
|---|---|---|---|
| `RTK_RAS` | defaults | `VAF >= 3.0` | defaults + `VAF >= 3.0` |
| `HRR`     | defaults | `ClinVar in (Pathogenic / Likely pathogenic)` | defaults + ClinVar restriction |
| `TP53`    | defaults | (none) | just the general defaults |
| other     | defaults | (none) | just the general defaults |

```{r oncoprint-filter-cleanup, include = FALSE}
ctdna_filter_scheme_delete("schemeStrict")
ctdna_filter_reset_general()
```

## 10.2 `ctdna_plot_alteration_grid(df, gene_sets, patient_data, response_col, indication_col, response_levels, scheme, filter, ..., stat, show_counts, ...)`

Stacked-bar grid: rows = indication, columns = gene set, x within each
panel = response category, fill = alt vs wt. Per-panel Fisher (or
chi-squared) p-value at top-left of each panel; `%alt (n_alt / n_total)`
label inside the alt segment (when ≥30% of the bar) or above the bar
(when <30%) so labels never overlap.

```{r alt-grid, fig.width = 11, fig.height = 6}
ctdna_plot_alteration_grid(sim$infinity_report,
  gene_sets = list(
    KRAS  = "KRAS",
    HRR   = c("BRCA1","BRCA2","ATM","CHEK2","PALB2"),
    RTK   = c("EGFR","ERBB2","MET","ALK","ROS1"),
    TP53  = "TP53"
  ),
  patient_data    = sim$clinical,
  response_col    = "RECIST",
  indication_col  = "Cancertype",
  scheme          = "three",                       # collapse x-axis RECIST values
  stat            = "fisher",
  filter          = list(functional_impact = c("High","Moderate"),
                          verbose = FALSE),
  title = "Combinatorial alteration status: response × indication × gene set"
)$plot
```

---

# 11. Theme, scales, palettes

The styling helpers can be applied to any custom ggplot.

**Visual encoding convention** (followed by every package plot):

| Feature | Aesthetic |
|---|---|
| RECIST | colour (always when shown) |
| Dose + RECIST shown together | shape (dose) |
| Dose alone | colour |

## 11.1 `ctdna_colors()` — RECIST palette

```{r colors}
ctdna_colors()              # both collapsed labels AND raw RECIST values
```

## 11.2 `ctdna_dose_palette(n)` — dose palette

```{r dose-pal}
ctdna_dose_palette(n = 3)
ctdna_dose_palette(n = 5)   # interpolates if more levels needed
```

## 11.3 `ctdna_theme(base_size)` and the scales

Apply the package theme + scales to a custom plot built from scratch.

```{r custom-recist, fig.height = 4}
base <- subset(ctdna_df, time_point == "Baseline")
base$RECIST <- ctdna_stratify_recist(base$RECIST, "three")

ggplot(base, aes(RECIST, methylTF, color = RECIST, fill = RECIST)) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA, linewidth = 0.7) +
  geom_jitter(width = 0.15, shape = 21, stroke = 0.3, alpha = 0.85) +
  scale_y_log10() +
  ctdna_scale_recist() +          # ggplot2 colour scale for RECIST
  ctdna_scale_recist_fill() +     # matching fill at lower alpha
  ctdna_theme(base_size = 11) +
  labs(title = "Custom plot — ctdna_theme + ctdna_scale_recist + ctdna_scale_recist_fill")
```

```{r custom-dose, fig.height = 3.5}
# Dose alone → colour + matching fill
ggplot(base, aes(dose, methylTF, color = dose, fill = dose)) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.15, shape = 21, stroke = 0.3, alpha = 0.85) +
  scale_y_log10() +
  ctdna_scale_dose() + ctdna_scale_dose_fill() + ctdna_theme() +
  labs(title = "Dose alone — ctdna_scale_dose + ctdna_scale_dose_fill")

# Dose + RECIST → RECIST=colour, dose=shape
ggplot(base, aes(dose, methylTF, color = RECIST, shape = dose)) +
  geom_jitter(width = 0.2, alpha = 0.85, size = 2.5) + scale_y_log10() +
  ctdna_scale_recist() + ctdna_scale_dose_shape() + ctdna_theme() +
  labs(title = "Dose + RECIST — ctdna_scale_dose_shape")
```

---

# 12. Publication-ready tables — `ctdna_table(x, title, subtitle, caption, row_names, highlight_diag, number_format, font_family, base_size)`

Renders any data frame, matrix, or `table` as a styled ggplot panel
with:

- title / subtitle / caption
- alternating row stripes
- bold underlined header
- top/bottom rules
- optional diagonal-cell tint (`highlight_diag = TRUE`)

```{r table-contingency, fig.height = 3}
tab <- as.data.frame(matrix(c(12, 3, 1, 9), 2, 2,
                            dimnames = list(MR = c("Response","NonResponse"),
                                            RECIST = c("CR/PR","SD/PD/NE"))))
ctdna_table(tab,
            title          = "MR × RECIST contingency",
            subtitle       = "n = 25; Cohen's kappa = 0.62",
            caption        = "MR computed at C5D1; ratio threshold = 0.5",
            highlight_diag = TRUE)
```

```{r table-summary, fig.height = 4}
ctdna_table(head(ctdna_summary(ctdna_df), 6),
            title    = "Per-subject summary",
            subtitle = "Baseline TF, best ratio, last TF",
            caption  = "First 6 subjects",
            base_size = 9)
```

---

# 13. End-to-end pipeline — `ctdna_run_pipeline(ctdna_df, expr_df, tmb_df, ihc_df, g74_df, g500_df, sim, stat_position, legend_position, scheme, plot_overrides, compose, verbose)`

Runs every standard deliverable in one call. Returns a named list of
plots (and optionally a patchwork composite via `compose = TRUE`).

```{r pipeline, fig.height = 4.5}
out <- ctdna_run_pipeline(
  ctdna_df = mm$ctdna,
  expr_df  = mm$expression,
  tmb_df   = mm$tmb,
  ihc_df   = mm$ihc,
  g74_df   = mm$genomic_74,
  sim      = sim,
  stat_position   = "subtitle",
  legend_position = "bottom",
  scheme   = "three",
  verbose  = FALSE
)
names(out)
out$baseline
out$mr_recist
```

Per-plot overrides:

```{r pipeline-overrides, eval = FALSE}
out <- ctdna_run_pipeline(
  ctdna_df = mm$ctdna,
  plot_overrides = list(
    plot_baseline = list(vars = c("methylTF","maxAF"), stat = "wilcox"),
    plot_vs_tmb   = list(panel = "500genes")
  )
)
```

---

# Appendix — every exported function (cross-check)

```{r appendix}
sort(getNamespaceExports("ctdnaTM"))
```

---

# Session info

```{r sysinfo}
Sys.info()
```

---

For any function, type `?function_name` for the full Rd help.
