---
title: "ctdnaTM: End-to-End Walkthrough (v0.75.10)"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ctdnaTM: End-to-End Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r logo, echo = FALSE, results = "asis"}
logo_path <- system.file("extdata", "logo.png", package = "ctdnaTM")
if (nzchar(logo_path) && file.exists(logo_path))
  cat(sprintf('<img src="%s" style="float:right; width:120px; margin-left:1em" alt="ctdnaTM logo"/>\n\n', logo_path))
```

```{r setup, include = FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 7, fig.height = 4, dpi = 100)
library(ctdnaTM)
library(ggplot2)
```

# End-to-end in one page

```{r data}
sim  <- ctdna_make_mock_study(n_patients = 50, seed = 22)
prep <- ctdna_prepare(sim, verbose = FALSE)
d    <- prep$ctdna[prep$ctdna$Visit_name == ctdna_opts("baseline"), ]
```

## D1: baseline characteristics

```{r d1}
ctdna_plot_baseline(d, group = "RECIST")
```

## D2: Dose x RECIST

```{r d2}
ctdna_plot_baseline_dose_recist(d)
```

## D3: per-subject reduction

```{r d3}
ctdna_plot_reduction(prep$ctdna)
```

## Unified `ctdna_boxplot()` (new in v0.75.10)

```{r boxplot}
ctdna_boxplot(d, x = "Dose", color = "RECIST", facet = NULL,
              compare_by = "color", method = "methyl")
```

For more, see the function-tour and core-deliverables docs.

# New in v0.75.10

## `indications` argument — one-line cohort restriction

Every user-facing plot function now takes an `indications` argument. Pass a
character vector of Indication values to restrict the plot to that cohort
without pre-filtering data:

```r
ctdna_plot_baseline(prep$ctdna, indications = c("NSCLC","mCRPC"))
ctdna_plot_reduction(prep$ctdna, indications = "mCRPC")
ctdna_plot_oncoprint(prep$variants,
                     gene_sets   = list(TSG = c("TP53","RB1","PTEN")),
                     indications = "NSCLC")
```

The lookup order for the Indication column: `Indication` -> `indication` ->
`Cancertype` -> `Cancer_Type`, first found wins in the frame itself; if not
present, a Patient_ID join against `prep$clinical` is attempted.

## `ctdna_concordance_oncoprint()` — ctDNA vs tissue paired oncoprint

Transposed oncoprint (genes in columns, patients in rows) with two stacked
tracks per patient: ctDNA on top, tissue underneath. Same glyphs and palette
as `ctdna_plot_oncoprint()`, plus the same annotation surface
(`patient_data` / `annotations` / `show_freq_bar` / `alterations` /
`sort_genes` / `subtitle` / `caption` / `legend_position` / `indications`).

```r
res <- ctdna_concordance_oncoprint(
  prep,
  tissue_df   = prep$dnaseq,             # or sim$dnaseq
  gene_sets   = list(TSG = c("TP53","RB1","PTEN"),
                     HRR = c("BRCA1","BRCA2","ATM","PALB2"),
                     Cancer_50 = cancer_50),
  visit       = "baseline",
  wrap        = "Indication",
  indications = c("mCRPC","NSCLC"),
  annotations = c("Dose","RECIST"),
  sort_genes  = "within_set",
  max_patients_per_panel = 20,
  title       = "ctDNA vs tissue concordance")

# res is a ggplot -- chain layers directly
res + ggplot2::theme(axis.text = ggplot2::element_text(size = 6))

# Auxiliary data via attributes
dims <- attr(res, "plot_dim")
ggplot2::ggsave("concordance.png", res,
                width = dims["width"], height = dims["height"],
                dpi = 120, limitsize = FALSE)
attr(res, "ctdna_matrix")   # patient x gene wide matrix
attr(res, "tissue_matrix")  # same for tissue
```

Requires the `patchwork` and `ggnewscale` packages
(`install.packages(c("patchwork","ggnewscale"))`).

## `cancer_50` — 50 pan-cancer genes

Ready-made character vector for the concordance oncoprint and other plots:

```r
head(cancer_50, 10)
# "TP53" "PIK3CA" "KRAS" "APC" "PTEN" "BRAF" "ARID1A" "KMT2D" "EGFR" "NRAS"
length(cancer_50)   # 50
```

## Multi-file dnaseq / rnaseq readers in `ctdna_prepare()`

When tissue variant calls live in per-sample TSVs across a directory tree,
you no longer need to pre-assemble them yourself. Point `ctdna_prepare()`
at the tree via a two-piece spec:

```r
prep <- ctdna_prepare(
  infinity_report = my_infinity_report,
  clinical        = my_clinical,
  dnaseq          = list(
    loc   = "/mnt/etc1/etc2/",
    regex = "*/variant/annotated/*.clinical.annotated.tsv"),
  verbose = TRUE)
```

Behavior:

- `loc` is a directory root, `regex` is a **glob pattern** relative to it.
  Dots are literal, `*` is any character except path separator.
- The **first `*` in the filename** (last path segment) becomes Sample_ID.
  For a match like `.../sampleA/.../sampleA.clinical.annotated.tsv` with
  pattern `*.clinical.annotated.tsv`, Sample_ID = `sampleA`.
- Each per-sample file is read, columns are canonicalized via
  `ctdna_opts()` keys (see below), and everything is concatenated.
- A progress bar shows per-file status.

You can also set the location and regex in `ctdna_opts()` and leave the arg
empty:

```r
ctdna_opts(dnaseq_loc   = "/mnt/etc1/etc2/",
           dnaseq_regex = "*/variant/annotated/*.clinical.annotated.tsv")
prep <- ctdna_prepare(infinity_report = ..., clinical = ...)  # picks up
```

## Column-name customization via `ctdna_opts()`

Since tissue-DNA output formats vary by vendor / pipeline, canonical column
names are user-configurable. Defaults (matching common WES pipelines):

| opts key                        | default          | required |
|---|---|:-:|
| `dnaseq_gene_col`               | `"Gene Symbol"`  | yes |
| `dnaseq_variant_type_col`       | `"Variant Type"` | yes |
| `dnaseq_vaf_col`                | `"Allelic Fraction"` | yes |
| `dnaseq_variant_effect_col`     | `"Variant Effect"` | no |
| `dnaseq_protein_variant_col`    | `"Protein Variant"` | no |
| `dnaseq_functional_class_col`   | `"Functional Class"` | no |
| `dnaseq_somatic_status_col`     | `"Somatic_status"` | no |
| `dnaseq_qc_filter_col`          | `"Filter"`       | no  |
| `dnaseq_min_vaf`                | `0`              | -   |

Set any of them with:

```r
ctdna_opts(dnaseq_gene_col  = "gene_symbol",
           dnaseq_vaf_col   = "VAF",
           dnaseq_min_vaf   = 0.05)
```

If `dnaseq_qc_filter_col` exists in the source files, only rows where that
column equals `"PASS"` are kept. If the column is absent, everything is
kept (assumed all PASS). `dnaseq_min_vaf` filters rows below the threshold
after normalizing to the 0-1 range.

Same set of keys exists for `rnaseq_*` (`rnaseq_loc`, `rnaseq_regex`,
`rnaseq_gene_col` = `"Gene Symbol"`, `rnaseq_value_col` = `"TPM"`,
`rnaseq_qc_filter_col` = `"Filter"`).
