---
title: "ctdnaTM — Core deliverables (v0.75.10)"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

```{r logo, echo = FALSE, results = "asis"}
logo_path <- system.file("extdata", "logo.png", package = "ctdnaTM")
if (nzchar(logo_path) && file.exists(logo_path))
  cat(sprintf('<img src="%s" style="float:right; width:120px; margin-left:1em" alt="ctdnaTM logo"/>\n\n', logo_path))
```

```{r setup, include = FALSE}
rm(list = ls())
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)
if (!requireNamespace("ctdnaTM", quietly = TRUE))
  install.packages(
    "/mnt/code/Hn/lib/ctdnaTM_0.43.8.tar.gz",
    repos = NULL, type = "source"
  )
library(ctdnaTM)
library(ggplot2)
```

# Overview

This document demonstrates the **canonical ctDNA deliverables** at
v0.75.10, with one section per plot function. The three baseline-and-
reduction deliverables (D1, D2, D3) plus the new unified function
**`ctdna_boxplot()`** are covered first; downstream deliverables
(longitudinal, MR–RECIST, etc.) follow.

For full API reference see `?ctdna_plot_baseline`, `?ctdna_plot_boxplot`,
etc. For the changelog see `NEWS.md`.

## Test data

All examples use the same mock cohort.

```{r data}
sim  <- ctdna_make_mock_study(n_patients = 200, seed = 22)
prep <- ctdna_prepare(sim, verbose = FALSE)

d_full <- prep$ctdna
d      <- d_full[d_full$Visit_name == ctdna_opts("baseline"), ]

cat("Baseline rows:", nrow(d), "\n")
cat("Longitudinal rows:", nrow(d_full), "\n")
cat("Doses:", paste(levels(d$Dose), collapse = ", "), "\n")
cat("RECIST:", paste(unique(d$RECIST), collapse = ", "), "\n")
```

---

# D1 — `ctdna_plot_baseline()`

Baseline ctDNA characteristics. One facet per TF method (when
`method = "all"`), boxes grouped by `group` (Dose or RECIST).

## D1 default: by RECIST

```{r d1_recist, fig.height = 4}
ctdna_plot_baseline(d, group = "RECIST")
```

## D1: by Dose, log scale

```{r d1_dose, fig.height = 4}
ctdna_plot_baseline(d, group = "Dose")
```

## D1: single method, custom ylim

```{r d1_single}
ctdna_plot_baseline(d, method = "methyl", group = "RECIST",
                     ylim = c(0.001, 1))
```

## D1: RECIST collapsed to two arms (R vs NR)

```{r d1_two}
ctdna_plot_baseline(d, group = "RECIST", scheme = "two")
```

**v0.75.10 highlights for D1:**

- `ylim` parameter added (passes to `coord_cartesian(ylim = ...)`).
- n-labels render INSIDE each panel at `y = -Inf, vjust = -0.4`;
  no longer collide with the next row in multi-method facet layouts.
- Caption reports overall p-value PER PANEL when multiple methods
  are facetted.

---

# D2 — `ctdna_plot_baseline_dose_recist()`

Baseline TF by Dose × RECIST. Dose on x, RECIST as colour, dodged
boxplots. Between-dose stats (pooling RECIST).

## D2 default: methyl, three-arm RECIST

```{r d2_default}
ctdna_plot_baseline_dose_recist(d)
```

## D2: max VAF

```{r d2_maxvaf}
ctdna_plot_baseline_dose_recist(d, method = "maxvaf")
```

## D2: all three methods stacked

```{r d2_all, fig.height = 9}
ctdna_plot_baseline_dose_recist(d, method = "all")
```

## D2: custom ylim and two-arm RECIST

```{r d2_two}
ctdna_plot_baseline_dose_recist(d, scheme = "two",
                                  ylim = c(0.001, 1))
```

**v0.75.10 highlights for D2:**

- `ylim` parameter added (passed through the multi-method recursion
  so all panels in the patchwork stack get the same clip).
- n-labels inside the panel.
- Bracket-builder fixes (linear-anchor at `v_data_max`, tighter
  `expand_top`) affect this function via the shared `.add_brackets`.

---

# D3 — `ctdna_plot_reduction()`

Per-subject reduction. Computes a derived metric via
`ctdna_metric_at()` and plots it as a Dose × RECIST boxplot.

## D3 default: ratio at best

```{r d3_default}
ctdna_plot_reduction(d_full)
```

## D3: percent change at last visit (linear scale)

```{r d3_pct_last}
ctdna_plot_reduction(d_full, metric = "pct_change", at = "last",
                     y_scale = "linear")
```

## D3: at a specific time-point

```{r d3_c2d1}
ctdna_plot_reduction(d_full, metric = "ratio", at = "C2D1")
```

## D3: tighter MR threshold

```{r d3_threshold}
ctdna_plot_reduction(d_full, mr_threshold = 0.3)
```

## D3: all three methods, stacked

```{r d3_all_methods, fig.height = 12}
ctdna_plot_reduction(d_full, method = "all")
```

**v0.75.10 highlights for D3:**

- `metric = "pct_change"`: `baseline_tf` floored at the LOQ
  (`ctdna_opts("display_floor")`) before dividing. Prevents
  pathological values like +49,900% from patients with near-LOQ
  baselines.
- All geoms (box, point, brackets, stats) use the same `d_capped`
  frame (clipped to `ylim`). With `clip = "off"`, this stops
  outliers from rendering above the panel.
- `ylab` defaults reflect the method label ("Max VAF",
  "Methylation TF", ...) instead of hardcoded "TF".
- `match.arg(metric)` moved above multi-method recursion (was
  failing on `method = "all"`).
- `ylim` is the canonical name; `y_clip` kept as deprecated alias.

---

# `ctdna_boxplot()` — unified function (NEW in v0.75.10)

Single configurable boxplot covering the structural space of
D1 + D2 + D3. Use it when you want to vary `x` / `color` / `facet` /
`compare_by` independently.

## Reaching D1 from `ctdna_boxplot()`

```{r boxplot_d1}
ctdna_boxplot(d, x = "RECIST")                          # D1 default
```

## Reaching D2 from `ctdna_boxplot()`

```{r boxplot_d2}
ctdna_boxplot(d, x = "Dose", color = "RECIST", facet = NULL,
              method = "methyl")
```

## Reaching D3-shape from `ctdna_boxplot()`

D3's per-subject metric isn't available as a column on the baseline
frame, but if you have any reduction-style column pre-computed (e.g.,
`best_pct_change` from `ctdna_prepare()`), `ctdna_boxplot()` handles
it natively:

```{r boxplot_d3}
ctdna_boxplot(d, x = "Dose", color = "RECIST", facet = NULL,
              value = "best_pct_change",
              y_scale = "linear", ylim = c(-100, 200))
```

## `compare_by` — what bracket comparisons do

Three modes:

| `compare_by` | What it compares | Brackets |
|---|---|---|
| `"x"` (default) | Pairwise between x groups, pooling colours | ONE stack per panel |
| `"color"` | Within each x level, pairwise between colour groups | ONE stack per x level (at dodged sub-positions) |
| `"x_within_color"` | Pairwise between x groups within each colour | Auto-facets by colour |

```{r compare_by_color}
# Compare RECIST groups WITHIN each Dose:
ctdna_boxplot(d, x = "Dose", color = "RECIST", facet = NULL,
              method = "methyl", compare_by = "color")
```

```{r compare_by_x_within_color, fig.height = 4}
# Compare Doses WITHIN each RECIST group (colour becomes a facet):
ctdna_boxplot(d, x = "Dose", color = "RECIST", facet = NULL,
              method = "methyl", compare_by = "x_within_color")
```

## `ctdna_boxplot()` validation guards

These reuse-of-column mistakes are rejected with a clear `stop()`:

```{r boxplot_guards, error = TRUE}
ctdna_boxplot(d, x = "Dose", color = "RECIST", facet = "Dose")
# Error: `facet` and `x` are both 'Dose'. Each facet would contain
# only one x value, leaving empty boxplots ...

ctdna_boxplot(d, x = "Dose", color = "Dose", facet = NULL)
# Error: `color` and `x` are both 'Dose'. The colour split would be
# redundant with the x-axis ...
```

---

# Other deliverables (unchanged in v0.75.10)

These deliverables exist and are unchanged in v0.75.10; their
documentation has not been updated since v0.38.0 doc baseline. See
the function help and the function tour Rmd for full details.

## D4 — Longitudinal TF trajectories

```{r d4, fig.height = 5}
ctdna_plot_longitudinal(d_full)
```

## D5 — Longitudinal by RECIST

```{r d5, fig.height = 4}
ctdna_plot_longitudinal_recist(d_full)
```

## D6 — MR–RECIST concordance

```{r d6, fig.height = 4}
ctdna_plot_mr_recist(d_full)
```

## D7 — Reduction vs % tumor change

```{r d7, fig.height = 4}
ctdna_plot_reduction_vs_tumor(d_full)
```

## D8 — Mutation vs methylation concordance

```{r d8, fig.height = 4}
ctdna_plot_mut_methyl(d_full)
```

## D9 — % change by dose and time

```{r d9, fig.height = 5}
ctdna_plot_pct_change_by_dose_time(d_full)
```

---

# Session info

```{r session, echo = FALSE}
sessionInfo()
```
