---
title: "ctdnaTM: end-to-end walkthrough"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{ctdnaTM: end-to-end walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                     fig.width = 7, fig.height = 4.5, dpi = 100,
                     message = FALSE, warning = FALSE)
```

One-page pass through `ctdnaTM`: prepare → QC → filter → plots → pipeline. For a deeper tour of every function see `vignette("ctdnaTM_function_tour")`. For the filter system's internals see `vignette("ctdnaTM_filter_internals")`.

# 1. Assemble the prep object

`ctdnaTM` builds every deliverable off a single **prep** object — a named list holding `$clinical`, `$samples`, `$variants`, and optionally `$assessments`, `$adrs`, and platform-specific frames. `ctdna_prepare()` assembles it from a Guardant Health Infinity report plus any ADaM frames.

```{r}
library(ctdnaTM)

# Real usage: replace with your infinity_report + ADaM frames.
sim  <- ctdna_make_mock_study(n_patients = 60, seed = 1)
prep <- ctdna_prepare(
  infinity_report = sim$infinity_report,
  adam            = list(adsl = sim$clinical),
  verbose         = FALSE)

class(prep)
setdiff(names(prep), "dictionary")
```

In v0.71+, `ctdna_prepare()` no longer applies sample QC by default — `prep$variants` comes back raw. This is deliberate: two users looking at the same prep will see the same rows.

# 2. Sample QC (idempotent)

```{r}
prep <- ctdna_sample_qc(prep, verbose = FALSE)
nrow(prep$variants)               # after QC
prep <- ctdna_sample_qc(prep, verbose = FALSE)
nrow(prep$variants)               # unchanged on rerun
```

# 3. Variant filtering

Single entry point with two mandatory flags:

| `apply` | `explain` | effect |
|:-:|:-:|:--|
| `TRUE`  | `TRUE`  | `prep$variants` filtered; annotated snapshot at `prep$filter_explanation$<sig>` |
| `TRUE`  | `FALSE` | `prep$variants` filtered; no snapshot |
| `FALSE` | `TRUE`  | `prep$variants` unchanged; snapshot only |
| `FALSE` | `FALSE` | error |

Live `prep$variants` never carries the `filtering_criteria` column — it lives only on the snapshot.

```{r}
prep <- ctdna_variant_filter(prep,
          filter_scheme = c("TSG_Tier1","TSG_Tier2","HRR14"),
          apply         = TRUE,
          explain       = TRUE,
          verbose       = FALSE)

# Row counts in the annotated snapshot
sig <- names(prep$filter_explanation)[1]
sig
head(sort(table(prep$filter_explanation[[sig]]$filtering_criteria,
                useNA = "ifany"), decreasing = TRUE), 6)
```

**Branch labels** are baked into every built-in scheme, so the annotation tells you *why* a row passed (or failed), not just *which* scheme:

- `PASS::TSG_Tier1(somatic_nonsyn), TSG_Tier2(somatic_LoF)` — passed both schemes, with the branch(es) that fired.
- `FAIL::TSG_Tier1, TSG_Tier2` — in both scopes, both failed.
- `NA` — row in no scheme's scope.

# 4. Built-in schemes

```{r}
# All four biomarker schemes drop ClinVar benign and CHIP unconditionally.
visualized_scheme(TSG_Tier2(), width = 76)
```

- `TSG_Tier1()` — most permissive on TP53 / RB1 / PTEN.
- `TSG_Tier2()` — intermediate on TP53 / RB1 / PTEN (old TSG body + new intra-chromosomal LGR).
- `TSG_Tier3()` — most restrictive on TP53 / RB1 / PTEN. No LGR, no fusion, no rare-syn carve-in.
- `HRR14()` — HRR biomarker on 14 HRR genes (old HRR body + new intra-genic LGR).
- `scheme_basic()` — GH Infinity "impactful alterations" ruleset, applied to all genes.

# 5. Build your own scheme

Two entry points:

```{r}
# Flat AND/OR case (positive criteria only, no negation)
my_short <- ctdna_create_scheme(
  name     = "my_short_tsg",
  genes    = c("TP53","RB1","PTEN"),
  criteria = c("somatic","truncating"),
  combine  = "all",
  overwrite = TRUE)

# Full grammar (with negation, nesting, named branches)
my_full <- create_filtering_scheme(
  rule(Gene = c("TP53","RB1","PTEN")),
  not(rule_clinvar_benign()),
  not(rule_ch()),
  branches = list(
    germline_PLP  = allOf(rule_germline(), rule_clinvar_path()),
    somatic_LoF   = allOf(rule_somatic(),  rule_truncating()),
    somatic_rare_missense = allOf(rule_somatic(),
                                  rule(Molecular_consequence = "missense"),
                                  rule_rare_001())),
  name = "my_full_tsg", overwrite = TRUE)
```

Both are usable with `ctdna_variant_filter(prep, "my_short_tsg", ...)`.

# 6. Landscape plots

```{r, fig.height = 4.5}
prep$adrs <- data.frame(
  Patient_ID = prep$clinical$Patient_ID,
  indication = rep(c("NSCLC","HNSCC","SCLC","mCRPC"),
                   length.out = nrow(prep$clinical)))

op <- ctdna_oncoprint(prep,
        gene_sets     = list(TSG   = c("TP53","RB1","PTEN"),
                             HRR14 = ctdna_gene_set("HRR14")),
        filter_scheme = "TSG_Tier1",
        engine        = "ggplot",
        title         = "Baseline landscape")
op$plot
```

```{r, fig.height = 3.5}
gr <- ctdna_alteration_grid(prep,
        gene_sets = list(TSG   = c("TP53","RB1","PTEN"),
                         HRR14 = ctdna_gene_set("HRR14")),
        scheme    = "three")
gr$plot
```

# 7. Whole pipeline in one call

`ctdna_pipeline()` runs every deliverable step whose input modalities are present. Access via `res$<key>$plot`.

```{r}
res <- ctdna_pipeline(prep,
         which     = c("oncoprint","alteration_grid"),
         scheme    = "three",
         gene_sets = list(TSG   = c("TP53","RB1","PTEN"),
                          HRR14 = ctdna_gene_set("HRR14")),
         engine    = "ggplot",
         verbose   = FALSE)
names(res)
```

# Where next

- `vignette("ctdnaTM_function_tour")` — every user-facing function with a runnable example.
- `vignette("ctdnaTM_filter_internals")` — how the filter engine works; how to extend the grammar.
- `?ctdna_rules_library` — the full rule-predicate catalog.
- `NEWS.md` — changelog.

# Session info

```{r}
sessionInfo()
```
