---
title: "ctdnaTM — Function tour (v0.75.10)"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

```{r logo, echo = FALSE, results = "asis"}
logo_path <- system.file("extdata", "logo.png", package = "ctdnaTM")
if (nzchar(logo_path) && file.exists(logo_path))
  cat(sprintf('<img src="%s" style="float:right; width:120px; margin-left:1em" alt="ctdnaTM logo"/>\n\n', logo_path))
```

```{r setup, include = FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 7, fig.height = 4, dpi = 100)
if (!requireNamespace("ctdnaTM", quietly = TRUE))
  install.packages(
    "/mnt/code/Hn/lib/ctdnaTM_0.43.8.tar.gz",
    repos = NULL, type = "source")
library(ctdnaTM)
library(ggplot2)
```

# About this document

Reference tour of every exported function in **ctdnaTM v0.75.10**.
For each function: signature, what it does, minimal example. For full
argument docs see the help page (`?function_name`).

```{r data_setup}
sim  <- ctdna_make_mock_study(n_patients = 60, seed = 22)
prep <- ctdna_prepare(sim, verbose = FALSE)
d_full <- prep$ctdna
d      <- d_full[d_full$Visit_name == ctdna_opts("baseline"), ]
```

---

# 1. Mock data & preparation

## `ctdna_make_mock_study()`

Generates a synthetic GH-Infinity-style multi-modal study bundle.
Returns a named list of frames.

```{r}
sim_small <- ctdna_make_mock_study(n_patients = 5, seed = 1)
names(sim_small)
```

## `ctdna_make_patient_data()`

Per-patient ADaM-style records (subset of `ctdna_make_mock_study()`'s
clinical frame).

```{r}
pat <- ctdna_make_patient_data(n_patients = 5, seed = 1)
str(pat, max.level = 1)
```

## `ctdna_prepare()`

**v0.75.10**: pure format conversion. No filtering applied — use
`ctdna_qc_filter()` afterwards for sample-level QC.

```{r}
# Mode (a): named list (the original v0.40.0 API)
prep_small <- ctdna_prepare(sim_small, verbose = FALSE)
names(prep_small)

# Mode (b): pass frames directly as named arguments
prep_alt <- ctdna_prepare(infinity_report      = sim_small$infinity_report,
                          tf_change            = sim_small$tf_change,
                          panel_74_response    = sim_small$panel_74_response,
                          panel_500_response   = sim_small$panel_500_response,
                          methylation_response = sim_small$methylation_response,
                          clinical             = sim_small$clinical,
                          verbose              = FALSE)
identical(names(prep_small), names(prep_alt))

# Mode (c): list-form spec with per-modality renames (v0.42.12+)
# prep <- ctdna_prepare(
#   infinity_report = list(data = my_inf,
#                          patient_id_col = "SUBJID",
#                          gene_col       = "MyGene"),
#   clinical        = list(data = my_clin,
#                          patient_id_col = "USUBJID")
# )
```

## `ctdna_qc_filter()`

**New in v0.75.10.** Apply sample-level QC AFTER `ctdna_prepare()`.
Reads `prep$samples` (built automatically by the GH Infinity
handler) and filters the longitudinal / genomic frames.

```{r}
# Default args are no-op:
prep_noop <- ctdna_qc_filter(prep_small, verbose = FALSE)
identical(prep_noop$samples, prep_small$samples)

# Real usage (would apply if final_df0-style data were present):
# prep_qc <- ctdna_qc_filter(prep_small,
#                             sample_status = "PASS",
#                             plasma_min    = 1,
#                             dna_ng_min    = 5)
```

## `ctdna_assemble_modalities()` / `ctdna_merge_modalities()`

Lower-level helpers used by `ctdna_prepare()`. See their help pages.

## `ctdna_cleanup()`

Applies the column-name and factor-level canonicalization step. Called
inside `ctdna_prepare()` when `cleanup = TRUE` (the default).

---

# 2. Core deliverable plots

## `ctdna_plot_baseline()` — D1

Baseline characteristics, one facet per TF method.

```{r d1}
ctdna_plot_baseline(d, group = "RECIST")
```

## `ctdna_plot_baseline_dose_recist()` — D2

Dose × RECIST dodged boxplots.

```{r d2}
ctdna_plot_baseline_dose_recist(d)
```

## `ctdna_plot_reduction()` — D3

Per-subject reduction (ratio / TF / pct_change at a given visit).

```{r d3}
ctdna_plot_reduction(d_full)
```

## `ctdna_boxplot()` — D1/D2/D3 unified (NEW v0.75.10)

Configurable x / colour / facet / compare_by.

```{r boxplot}
ctdna_boxplot(d, x = "RECIST")                              # D1-shape
ctdna_boxplot(d, x = "Dose", color = "RECIST", facet = NULL,
              method = "methyl")                            # D2-shape
ctdna_boxplot(d, x = "Dose", color = "RECIST", facet = NULL,
              compare_by = "color", method = "methyl")      # new mode
```

## `ctdna_plot_longitudinal()` — D4

Longitudinal TF trajectories.

```{r d4}
ctdna_plot_longitudinal(d_full)
```

## `ctdna_plot_longitudinal_recist()` — D5

Longitudinal by RECIST response.

```{r d5}
ctdna_plot_longitudinal_recist(d_full)
```

## `ctdna_plot_mr_recist()` — D6

MR vs RECIST concordance heatmap / scatter.

```{r d6}
ctdna_plot_mr_recist(d_full)
```

## `ctdna_plot_reduction_vs_tumor()` — D7

Per-subject reduction vs % tumor change.

```{r d7}
ctdna_plot_reduction_vs_tumor(d_full)
```

## `ctdna_plot_mut_methyl()` — D8

Mutation-based vs methylation-based TF comparison.

```{r d8}
ctdna_plot_mut_methyl(d_full)
```

## `ctdna_plot_pct_change_by_dose_time()` — D9

% change in TF by dose and time-point.

```{r d9}
ctdna_plot_pct_change_by_dose_time(d_full)
```

---

# 3. Other plot functions

## `ctdna_plot_oncoprint()`

Standard oncoprint over baseline alterations.

```{r oncoprint}
# Requires the alterations table; uses ComplexHeatmap if installed
try(ctdna_plot_oncoprint(prep$genomic), silent = TRUE)
```

## `ctdna_plot_alteration_grid()`

Gene × patient grid.

## `ctdna_plot_vs_expression()` / `ctdna_plot_vs_ihc()` / `ctdna_plot_vs_mutation()` / `ctdna_plot_vs_tmb()`

ctDNA vs an external modality (expression, IHC, mutation status, TMB).

---

# 4. Statistics

## `ctdna_pairwise_ranksum()`

Pairwise Wilcoxon between groups.

```{r}
res <- ctdna_pairwise_ranksum(d, group_col = "Dose",
                                value_col = "methylTF")
res
```

## `ctdna_cor_test()`

Correlation test wrapper.

## `ctdna_paired_test()`

Paired t / signed-rank tests between two time-points.

## `ctdna_concordance_test()`

Cohen's kappa for MR vs RECIST.

## `ctdna_cohen_kappa()` / `ctdna_run_group_test()`

Used internally by the test wrappers; exported for direct use.

## `ctdna_pval_label()`

Format a p-value into the standard star/significance string.

```{r}
ctdna_pval_label(0.00021)
ctdna_pval_label(0.04)
ctdna_pval_label(0.51)
```

---

# 5. Metric computation

## `ctdna_metric_at()`

Compute the per-subject metric at a given visit (used internally by
D3 and others, useful directly).

```{r}
m <- ctdna_metric_at(d_full, metric = "ratio", at = "best",
                       tf_col = "methylTF")
head(m, 3)
```

## `ctdna_compute_mr()`

Classify subjects as Molecular Responders / NonResponders.

## `ctdna_ratio()` / `ctdna_floor_tf()`

Helper transforms used pre-plot.

```{r}
ctdna_floor_tf(c(0, 0.0005, 0.5, NA))     # values < LOQ floored to LOQ
```

## `ctdna_to_log2tpm()`

Expression conversion helper.

## `ctdna_stratify_recist()`

Collapse RECIST into 2/3/4/5/6-arm schemes.

```{r}
ctdna_stratify_recist(c("CR", "PR", "SD", "PD", "NE"),
                       recist_grouping = "three")
```

---

# 6. Filter system

See the dedicated **filter system walkthrough** Rmd for the full rule
language. Brief summary here:

## `ctdna_filter_apply()`

The unified entry point. Two stages (sample QC + variant filter).

```{r}
# QC only (no variant filtering)
df_qc <- ctdna_filter_apply(prep$genomic, filter_scheme = NULL,
                              sample_qc_enable = TRUE)
nrow(df_qc)
```

## Schemes: `scheme_basic()`, `scheme_HRR()`, `scheme_TSG()`

Pre-defined rule sets.

```{r}
df_basic <- ctdna_filter_apply(prep$genomic,
                                 filter_scheme = scheme_basic())
nrow(df_basic)
```

## Combinators: `allOf()`, `anyOf()`, `"not"()`, `rule()`

Build composite rules.

```{r}
mc_rule <- allOf(rule_clinvar_path(), rule_truncating())
```

## Rule library: 80+ predicates

Listed in NAMESPACE — common groups:

- **Variant type**: `rule_is_SNV()`, `rule_is_Indel()`, `rule_is_CNV()`,
  `rule_is_Fusion()`, `rule_is_LGR()`
- **Consequence**: `rule_missense()`, `rule_nonsense()`, `rule_frameshift()`,
  `rule_synonymous()`, `rule_splice_donor()`, etc.
- **ClinVar**: `rule_clinvar_path()`, `rule_clinvar_benign()`,
  `rule_clinvar_vus()`, `rule_clinvar_conflict()`
- **Germline / somatic**: `rule_germline()`, `rule_somatic()`,
  `rule_ch()` (clonal haematopoiesis)
- **Frequency**: `rule_common_001()`, `rule_rare_001()`, `rule_rare_0001()`
- **Gene sets**: `rule_genes_HRR14()`, `rule_genes_TSG()`,
  `rule_genes_PI3K()`, `rule_genes_RTK()`, `rule_genes_MMR()`,
  `rule_genes_Cell_Cycle()`, `rule_genes_TP53_pathway()`
- **Indication**: `rule_ind_BRCA()`, `rule_ind_NSCLC()`, `rule_ind_SCLC()`,
  `rule_ind_CRC()`, `rule_ind_HNSCC()`, `rule_ind_GBM()`,
  `rule_ind_mCRPC()`
- **CNV/LGR specifics**: `rule_cnv_amp()`, `rule_cnv_focal_amp()`,
  `rule_cnv_homozyg_del()`, `rule_lgr_deletion()`,
  `rule_lgr_tandem_duplication()`, `rule_is_LGR_deleterious()`,
  `rule_amp_thresh()`, `rule_biallelic()`, `rule_reversion()`,
  `rule_BRCA2_K3326()`
- **Misc**: `rule_promoter()`, `rule_deleterious()`,
  `rule_truncating()`, `rule_sample_fail()`, `rule_sample_success()`

## Scheme management

- `ctdna_filter_schemes_list()`, `ctdna_filter_scheme_get()`,
  `ctdna_filter_scheme_save()`, `ctdna_filter_scheme_load()`,
  `ctdna_filter_scheme_delete()`, `ctdna_filter_scheme_show()`,
  `create_filtering_scheme()`, `visualized_scheme()`,
  `ctdna_test_scheme()`.

## Filter options

- `ctdna_filter_opts()`, `ctdna_filter_reset_general()`.

---

# 7. Annotation and integration

## `ctdna_annotate_population_freq()`

Annotate alterations with gnomAD population frequencies.

## `ctdna_load_popfreq_cache()` / `ctdna_save_popfreq_cache()`

Cache management for the population-frequency annotation.

## `ctdna_gene_set()`

Look up a named gene set.

```{r}
ctdna_gene_set("HRR14")
```

## `ctdna_check_for_updates()`

Check whether a newer ctdnaTM version is available.

---

# 8. Options and themes

## `ctdna_opts()`

Query / set library-wide options.

```{r}
ctdna_opts("baseline")              # baseline visit name
ctdna_opts("display_floor")         # LOQ
ctdna_opts("show_stats")            # default show_stats
```

## `ctdna_opts_choices()`

Show allowed values for option keys.

## `ctdna_opts_save()` / `ctdna_opts_load()`

Persist opts to disk.

## `ctdna_theme()` / `ctdna_colors()`

ggplot theme and palette.

## `ctdna_scale_recist()` / `ctdna_scale_recist_fill()`

RECIST colour scales.

## `ctdna_scale_dose()` / `ctdna_scale_dose_fill()` / `ctdna_scale_dose_shape()`

Dose colour / shape scales.

## `ctdna_dose_palette()`

Raw palette (named vector).

```{r}
ctdna_dose_palette()
```

---

# 9. Pipeline and method registry

## `ctdna_pipeline()` / `ctdna_run_pipeline()` / `ctdna_pipeline_steps()`

Composable pipeline runner.

## `ctdna_register_method()` / `ctdna_methods()` / `ctdna_alteration_types()`

Extend the TF-method registry (e.g. add a vendor-specific column).

```{r}
ctdna_methods(domain = "tf")
```

## `ctdna_platforms()`

List registered vendor platforms.

```{r}
ctdna_platforms()
```

---

# 10. Utilities

## `ctdna_help()`

Open the package help index.

## `ctdna_logo()`

Path to the logo file.

## `ctdna_summary()`

One-line summary of a prepared bundle.

## `ctdna_table()`

Helper to produce a clinical-style summary table.

## `ctdna_coerce()`

Convert a vector to a canonical factor (Dose, RECIST, etc.).

```{r}
ctdna_coerce(c("Low", "High", "Mid"), kind = "Dose")
```

## `clinical_opts()`

Options for the clinical-modality assembler.

---



# New in v0.75.10

## `ctdna_concordance_oncoprint()`

Transposed oncoprint pairing ctDNA and tissue calls per patient. Full doc:
`?ctdna_concordance_oncoprint`. Every argument surfaces via tab-autocomplete.

Signature (all args after `tissue_df` are optional):

```r
ctdna_concordance_oncoprint(
    prep,
    tissue_df,
    gene_sets              = NULL,
    filter_scheme          = NULL,
    visit                  = "baseline",
    wrap                   = NULL,
    indications            = NULL,
    patient_data           = NULL,
    annotations            = NULL,
    show_freq_bar          = TRUE,
    alterations            = NULL,
    sort_genes             = c("global","within_set","none"),
    max_patients_per_panel = 25,
    show_patient_names     = NA,
    sort_by                = c("combined_burden","ctdna_burden",
                                "tissue_burden","patient"),
    title                  = "ctDNA vs tissue concordance",
    subtitle               = NULL,
    caption                = NULL,
    legend_position        = "right")
```

Return: a `ggplot` (class `ctdna_concordance_oncoprint / patchwork / gg /
ggplot`) with these attributes attached — `ctdna_matrix`, `tissue_matrix`,
`plot_dim`, `patients`, `genes`, `long`, `n_dropped_by_topN`. Access via
`attr(res, "<name>")`.

## `cancer_50`

Pan-cancer gene set (character vector length 50). `?cancer_50` for the
composition (solid-tumor drivers, DDR/HRR, heme, RTKs, chromatin).

## `indications` argument on every plot function

Every user-facing plot function accepts `indications = NULL`. When set to a
character vector of Indication values, the input data.frame is restricted
to those patients before any other processing. Full list of functions with
the new arg (16 total, unchanged behavior elsewhere):

- `ctdna_plot_baseline`, `ctdna_plot_baseline_dose_recist`
- `ctdna_plot_reduction`, `ctdna_plot_mr_recist`
- `ctdna_plot_longitudinal`, `ctdna_plot_longitudinal_recist`
- `ctdna_plot_reduction_vs_tumor`, `ctdna_plot_mut_methyl`
- `ctdna_plot_pct_change_by_dose_time`, `ctdna_boxplot`
- `ctdna_plot_vs_expression`, `ctdna_plot_vs_mutation`
- `ctdna_plot_vs_tmb`, `ctdna_plot_vs_ihc`
- `ctdna_plot_alteration_grid`, `ctdna_plot_oncoprint`

Example:

```r
ctdna_plot_baseline(prep$ctdna, indications = "NSCLC")
ctdna_plot_oncoprint(prep$variants,
                     gene_sets   = list(TSG = c("TP53","RB1","PTEN")),
                     indications = c("mCRPC","NSCLC"))
```

## Multi-file readers in `ctdna_prepare()`

`dnaseq` and `rnaseq` slots now accept a `list(loc = ..., regex = ...)`
spec instead of a preassembled data.frame. The reader walks the tree,
matches files by glob, extracts Sample_ID from the filename wildcard,
canonicalizes columns via `ctdna_opts()`, and concatenates. Progress bar
via `utils::txtProgressBar`.

See `?ctdna_prepare` and the "Multi-file dnaseq / rnaseq readers" section
of the walkthrough vignette.

## New `ctdna_opts()` keys

`dnaseq_loc`, `dnaseq_regex`, `dnaseq_gene_col`, `dnaseq_variant_type_col`,
`dnaseq_vaf_col`, `dnaseq_variant_effect_col`, `dnaseq_protein_variant_col`,
`dnaseq_functional_class_col`, `dnaseq_somatic_status_col`,
`dnaseq_qc_filter_col`, `dnaseq_min_vaf`. Same set with `rnaseq_` prefix
(`rnaseq_gene_col`, `rnaseq_value_col`, `rnaseq_qc_filter_col`,
`rnaseq_loc`, `rnaseq_regex`).

# Session info

```{r session, echo = FALSE}
sessionInfo()
```
