---
title: "ctdnaTM filter system: internals"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ctdnaTM filter system: internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

How the filter system in `ctdnaTM` is put together — object shapes, evaluator, engine, semantics, predicate library, and how to extend it. Read `vignette("ctdnaTM_walkthrough")` first for the user-facing view.

# 1. Overview

The filter system has three tiers, from bottom to top:

1. **Rules** — atomic predicates and their combinators. `rule()`, `rule_*()`, `allOf()`, `anyOf()`, `not()`.
2. **Schemes** — named packages of rules with a specific structure (scope + gate + optional branches). Built via `create_filtering_scheme()` or `ctdna_create_scheme()`.
3. **Engine** — the runtime that applies a scheme (or several) to a variants frame. Entry point: `ctdna_variant_filter()`.

The engine consults **two catalogs** for named lookups:

- The **scheme catalog**, populated at package load from `.ctdna_builtin_schemes()` and extended by `ctdna_filter_scheme_save()`.
- The **rule library** (all `rule_*()` predicates), assembled at package load from every function in `R/rules_library.R`.

# 2. Object shapes

Every rule / scheme is an S3 object with a specific structure. Understanding these shapes makes it easy to introspect, debug, and extend the system.

## 2.1 `ctdna_rule` — atomic predicate

```{r}
library(ctdnaTM)
r <- rule(Gene = c("TP53","BRCA1"))
class(r)
str(r, max.level = 1)
```

Structure:

- `type = "atom"` — leaf node.
- `columns = c("Gene")` — the column(s) the rule reads from the variants frame.
- `values` — a list of expected values (or a predicate function).
- `label = "Gene ∈ {TP53, BRCA1}"` — human-readable summary used by `visualized_scheme()`.

An atomic rule evaluates row-by-row: returns a logical vector the same length as `nrow(variants)`, `TRUE` where the row matches.

## 2.2 Combinators — `type = "and" / "or" / "not"`

```{r}
r_and <- allOf(rule(Gene = "TP53"), rule_somatic(), rule_truncating())
r_or  <- anyOf(rule_frameshift(), rule_nonsense())
r_neg <- not(rule_ch())
c(r_and$type, r_or$type, r_neg$type)
```

Structure:

- `type` = one of `"and"`, `"or"`, `"not"`.
- `children` = list of rules (one child for `not`, ≥1 for `and`/`or`).
- `label` = auto-composed from children's labels.

Aliases: `allOf` == `alt_all`, `anyOf` == `alt_any`. There's also `alt_any_n(rules, n)` for "at least n children match" — internally an `or`-like node with an integer threshold.

## 2.3 `ctdna_filtering_scheme` — top-level scheme

```{r}
str(TSG_Tier2(), max.level = 1)
```

Structure:

- `name`, `description`, `frozen`.
- `scope` — a rule (usually a gene list) that determines *which variants this scheme is responsible for*. A variant outside every scheme's scope becomes `NA` in `filtering_criteria`.
- `gates` — a list of rules that are AND-combined with the branches. Typically negation rules like `not(rule_ch())`, `not(rule_clinvar_benign())`.
- `branches` — a **named** list of rules. A variant PASSes the scheme when it's in scope, passes all gates, AND matches at least one branch. The name of the matching branch appears in `filtering_criteria` as `PASS::<scheme>(<branch>)`.

If `branches` is `NULL`, the scheme is a simple `scope + gate` predicate with no PASS/FAIL breakdown per branch — the label reduces to `PASS::<scheme>` or `FAIL::<scheme>`.

# 3. The evaluator

Function: `.eval_rule(rule, df)` (internal). Recursive walker:

```
.eval_rule(rule, df):
  switch (rule$type)
    "atom"  ->  match df[[rule$columns]] against rule$values
    "and"   ->  Reduce("&", eval on each child)
    "or"    ->  Reduce("|", eval on each child)
    "not"   ->  !eval(child)
```

Every call returns a logical vector. `NA` handling: unknown values propagate as `NA`, but the final engine converts `NA` to `FALSE` before applying (so unknowns cannot silently PASS).

You can invoke the evaluator directly for debugging:

```{r}
sim <- ctdna_make_mock_study(n_patients = 30, seed = 1)
prep <- ctdna_prepare(sim$infinity_report,
                      adam = list(adsl = sim$clinical), verbose = FALSE)
prep <- ctdna_sample_qc(prep, verbose = FALSE)

# Direct evaluation on the variants frame
hit <- ctdnaTM:::.eval_rule(rule(Gene = c("TP53","RB1","PTEN")),
                                   prep$variants)
sum(hit)   # variants in scope
```

# 4. The filter engine

Function: `ctdna_variant_filter(prep, filter_scheme, apply, explain, unmatched, verbose)`.

Pipeline (per scheme, in the vector `filter_scheme`):

1. Resolve `filter_scheme` — if it's a character name, look it up in the catalog; if it's a `ctdna_filtering_scheme` object, use directly.
2. Evaluate the scheme against `prep$variants`:
   - Compute `in_scope <- eval(scope)`.
   - Compute `gate_pass <- Reduce("&", lapply(gates, eval))`.
   - For each branch: compute `branch_pass_<name>`.
   - `PASS <- in_scope & gate_pass & Reduce("|", branch_passes)`.
3. Build the per-row `filtering_criteria` label:
   - If not in scope of any active scheme → `NA`.
   - If in scope but PASS false → `FAIL::<scheme1>, <scheme2>, ...`.
   - If in scope and PASS true → `PASS::<scheme1>(<branch>), <scheme2>(<branch>)`.
4. Attach the annotated snapshot to `prep$filter_explanation$<sig>` if `explain = TRUE`.
5. Drop rows where PASS false (if `apply = TRUE`). Rows where all scheme labels are `NA` are dropped when `unmatched = "drop"`, kept when `unmatched = "keep"`.

Multiple schemes are OR-combined at the PASS level — a row passing any scheme survives.

# 5. Semantics reference

## 5.1 Truth table for a single scheme

| in scope? | gate pass? | any branch pass? | Label | Row fate (apply=TRUE) |
|:-:|:-:|:-:|:--|:-:|
| no  | –    | –    | `NA`                                | drop (unmatched=drop) / keep (unmatched=keep) |
| yes | no   | –    | `FAIL::<scheme>`                    | drop |
| yes | yes  | no   | `FAIL::<scheme>`                    | drop |
| yes | yes  | yes  | `PASS::<scheme>(<matched branch>)`  | keep |

## 5.2 Multiple schemes

Labels are concatenated in scheme order. A row that's in scope of two schemes and passes both:

```
PASS::TSG_Tier2(somatic_LoF), HRR14(deleterious_LGR)
```

A row in scope of both but failing both:

```
FAIL::TSG_Tier2, HRR14
```

A row in scope of one but not the other:

```
PASS::TSG_Tier2(somatic_LoF), HRR14::out_of_scope
```

## 5.3 `unmatched`

Only affects rows where every scheme label is `NA` (i.e. not in scope of any scheme).

- `"drop"` (default when `apply=TRUE`) — remove them.
- `"keep"` — leave them in `prep$variants` untouched.

## 5.4 Idempotence

Sample QC removes samples and cascades to variants. Variant filtering removes variants only (samples unaffected). Both operations are safe to rerun — a second run has no additional effect provided the input is the same.

# 6. Built-in schemes

## 6.1 `TSG_Tier1`

- **Scope:** `Gene ∈ {TP53, RB1, PTEN}`.
- **Gates:** `not(rule_ch()) AND not(rule_clinvar_benign())`.
- **Branches:**
  - `somatic_nonsyn` — any impactful somatic alteration.
  - `germline_PLP`   — germline pathogenic / likely-pathogenic.
  - `deleterious_LGR` — a deleterious LGR (large genomic rearrangement).

Broadest of the three tiers.

## 6.2 `TSG_Tier2`

- Same scope + gates as Tier 1.
- **Branches** tightened:
  - `somatic_LoF` — somatic AND loss-of-function (truncating / frameshift / splice-donor / splice-acceptor / start-lost / stop-lost).
  - `germline_PLP`, `deleterious_LGR` as in Tier 1.

## 6.3 `TSG_Tier3`

- Same scope + gates.
- **Branches** stricter still:
  - `germline_PLP`.
  - `somatic_LoF`.
  - `somatic_ClinVar_PLP_missense_inframe` — ClinVar-P/LP missense or inframe indel.
  - `LOH_homozyg_del` — copy-loss events.
- **No LGR, fusion, or rare-synonymous branches.**

## 6.4 `HRR14`

- **Scope:** the 14 canonical HRR genes (BRCA1/2, ATM, PALB2, CHEK1/2, RAD51*, FANC*, BRIP1, BARD1, BLM, MRE11).
- **Gates:** as in the TSG schemes plus `not(rule_BRCA2_K3326())` (the benign K3326\* polymorphism).
- **Branches:** three PASS branches (somatic_nonsyn, germline_PLP, deleterious_LGR) plus an **intra-genic LGR** branch (`Gene == Fusion_gene_b`) — LGR events entirely within one HRR gene.

## 6.5 `scheme_basic`

- **Scope:** everything on the panel.
- **Gates:** minimal (drop CH and ClinVar-benign).
- **Branches:** one — any impactful alteration (SNV, indel, CNV, LGR, or fusion).

Useful as a broad screening filter.

# 7. Adding a new built-in scheme

Two ways to add a scheme so it's discoverable by `filter_scheme = "<name>"` lookup.

## 7.1 Session-scoped (fast, ephemeral)

`create_filtering_scheme(..., overwrite = TRUE)` registers the scheme in the current R session:

```{r, eval = FALSE}
create_filtering_scheme(
  rule(Gene = c("KRAS","BRAF","EGFR")),
  not(rule_ch()),
  branches = list(hotspot = rule_missense()),
  name = "driver_hotspots",
  overwrite = TRUE)

# Now usable by name
prep <- ctdna_variant_filter(prep, "driver_hotspots",
                             apply = TRUE, explain = TRUE)
```

## 7.2 Persistent (survives R restart)

Save to disk:

```{r, eval = FALSE}
ctdna_filter_scheme_save("driver_hotspots",
                         path = "~/.ctdnaTM/driver_hotspots.rds")

# Later session
ctdna_filter_scheme_load("driver_hotspots",
                         path = "~/.ctdnaTM/driver_hotspots.rds")
```

## 7.3 Package-scoped (contributor level)

Add a constructor function to `R/rules_library.R` (or a similar file), then include it in `.ctdna_builtin_schemes()`:

```r
.ctdna_builtin_schemes <- function() {
  list(
    TSG_Tier1  = TSG_Tier1_scheme(),
    TSG_Tier2  = TSG_Tier2_scheme(),
    TSG_Tier3  = TSG_Tier3_scheme(),
    HRR14      = HRR14_scheme(),
    scheme_basic = scheme_basic_scheme(),
    driver_hotspots = driver_hotspots_scheme())   # new
}
```

# 8. Extending the predicate library

New predicates are plain functions that return a `ctdna_rule` object. Minimal skeleton:

```{r, eval = FALSE}
rule_vaf_gt_5 <- function() {
  rule(VAF_percentage = function(x) x > 5,
       label = "VAF > 5%")
}
```

Any function returning a `ctdna_rule` can be composed with the standard combinators. Ship it via the same channel as any other exported function (roxygen doc + `@export`).

Tip: use `label = "..."` explicitly if the auto-generated label is ugly — this string appears in `visualized_scheme()` output and helps debugging.

# 9. Debugging

## 9.1 `visualized_scheme(scheme)`

Prints the full tree of a scheme with human-readable labels:

```{r}
visualized_scheme(HRR14())
```

## 9.2 `ctdna_test_scheme(prep, scheme)`

Dry-runs a scheme against a prep. Reports counts at each level (in scope, gate pass, per-branch pass, final PASS).

```{r}
head(ctdna_test_scheme("TSG_Tier2", df = prep$variants))
```

## 9.3 The `filter_explanation` snapshot

When `ctdna_variant_filter(..., explain = TRUE)` runs, it stores an annotated copy of `prep$variants` with two additional columns:

- `filtering_criteria` — the `PASS::...` / `FAIL::...` / `NA` label per row.
- `filter_pass` — a boolean summarising PASS/FAIL/NA.

```{r}
prep <- ctdna_variant_filter(prep,
                             filter_scheme = c("TSG_Tier2","HRR14"),
                             apply = FALSE, explain = TRUE,
                             verbose = FALSE)
sig <- names(prep$filter_explanation)[1]
head(sort(table(prep$filter_explanation[[sig]]$filtering_criteria,
                useNA = "ifany"), decreasing = TRUE), 8)
```

## 9.4 Common pitfalls

**Rule matches nothing.** Usually a column-name mismatch. Check `names(prep$variants)` and confirm the rule's `columns` attribute:

```{r, eval = FALSE}
r <- rule(Gene = c("TP53","RB1"))
r$columns    # what the rule reads
```

**A branch never fires.** Check the branch's own evaluator output:

```{r, eval = FALSE}
b <- HRR14()$branches$somatic_nonsyn
sum(ctdnaTM:::.eval_rule(b, prep$variants))
```

**Every variant is `NA` in filtering_criteria.** No scheme claimed scope — either the gene list doesn't overlap your panel, or the scope column names don't match `prep$variants`.

**Filter drops too many.** Set `apply = FALSE, explain = TRUE` and inspect the label distribution. Look for `FAIL::...` rows and check which gate is firing.

# 10. Semantics summary

| Concept | Where it lives | How to change |
|---|---|---|
| Atomic rule    | `ctdna_rule` object                                    | `rule()`, `rule_*()` functions |
| Combinator     | `ctdna_rule` object, `type = "and"/"or"/"not"`         | `allOf`, `anyOf`, `not` |
| Scheme         | `ctdna_filtering_scheme` object                        | `create_filtering_scheme()`, `ctdna_create_scheme()` |
| Catalog        | in-memory registry populated at package load           | `ctdna_filter_scheme_save`/`load`/`delete` |
| Engine         | `ctdna_variant_filter()`                               | not user-configurable; behavior tuned via arguments |
| Semantics      | evaluator + engine, described above                    | contributor-level change |

# Session info

```{r}
sessionInfo()
```
