---
title: "ctdnaTM function tour"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ctdnaTM function tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5, dpi = 100,
                      message = FALSE, warning = FALSE)
```

Every exported function in `ctdnaTM`, grouped by purpose, each with a runnable example on mock data. For a linear tutorial see `vignette("ctdnaTM_walkthrough")`; for filter-engine internals see `vignette("ctdnaTM_filter_internals")`.

# 1. Setup & mock data

```{r}
library(ctdnaTM)
sim  <- ctdna_make_mock_study(n_patients = 60, seed = 7)
# oncoprint's wrap = "Indication" needs a lower-case `indication` column;
# mock ships Cancertype so we copy it over.
sim$clinical$indication <- sim$clinical$Cancertype

prep <- ctdna_prepare(
  infinity_report = sim$infinity_report,
  adam            = list(adsl = sim$clinical, adtr = sim$adtr),
  verbose         = FALSE)
prep <- ctdna_sample_qc(prep, verbose = FALSE)
setdiff(names(prep), "dictionary")
```

`ctdna_make_mock_study(n_patients, seed, ...)` returns a named list with an Infinity report, clinical/adtr frames, and orthogonal-platform frames (`ihc`, `dnaseq`, `rnaseq`) — enough to exercise every function without a real dataset.

# 2. Data assembly

## 2.1 `ctdna_prepare()`

The entry point. Accepts a Guardant Infinity report plus any ADaM frames, produces a `ctdna_prep` list with `$clinical / $samples / $variants / $assessments / $dictionary / $qc_removed`.

```{r, eval = FALSE}
prep <- ctdna_prepare(
  infinity_report = my_report,                                    # data.frame or list()-spec
  adam            = list(adsl = adsl, adrs = adrs, adtr = adtr),  # union across frames
  clinical        = NULL,                                         # alternative to adam=
  tf_change       = NULL,                                         # optional per-sample TF change frame
  panel_74        = NULL,                                         # optional 74-gene panel frame
  panel_500       = NULL,                                         # optional 500-gene panel frame
  methylation     = NULL,                                         # optional methylation frame
  verbose         = TRUE)
```

Key behavior:

- **Verbatim variant construction.** `prep$variants` mirrors your Infinity report row-for-row, plus a `Baseline` boolean flag.
- **Union merge for clinical.** Every frame in `adam` contributes patient IDs to `prep$clinical`. First non-NA value wins per column, in list order.
- **No auto-QC.** Sample QC happens only when *you* call `ctdna_sample_qc()`.
- **Any modality passed as a list-spec** — e.g. `infinity_report = list(data = df, patient_id_col = "SUBJID", maxvaf = "VAF_max")` — gets its rename hints applied so canonical names line up.

## 2.2 `ctdna_prep_add()`

Attach any modality frame to an already-built prep. Useful for adding IHC, RNAseq, or a second sequencing panel after the fact.

```{r}
prep <- ctdna_prep_add(prep, ihc     = sim$ihc)
prep <- ctdna_prep_add(prep, dnaseq  = sim$dnaseq)
# sim$rnaseq is a nested list (counts / tpm / sample_metadata);
# pass the tpm matrix explicitly:
prep <- ctdna_prep_add(prep, expression = sim$rnaseq$tpm)
setdiff(names(prep), "dictionary")
```

## 2.3 `ctdna_platforms()`

List every recognised platform-alias mapping — what canonical names ctdnaTM knows and what columns of each vendor's export it recognises.

```{r}
head(names(ctdna_platforms()))
```

# 3. QC & variant filtering

## 3.1 `ctdna_sample_qc()`

Applies sample-level QC criteria, cascades to variants tied to failed samples, archives removed rows.

```{r}
prep_qc <- ctdna_sample_qc(prep,
                           sample_status = c("SUCCESS"),
                           plasma_min    = 0,
                           cfdna_min     = 0,
                           verbose       = FALSE)
sapply(prep_qc$qc_removed, function(x) if (is.null(x)) 0L else nrow(x))
```

Idempotent — running it twice does not double-cascade.

## 3.2 `ctdna_variant_filter()`

Applies one or more schemes to `prep$variants`. `apply` and `explain` are mandatory.

```{r}
prep <- ctdna_variant_filter(prep,
                             filter_scheme = c("TSG_Tier1", "HRR14"),
                             apply         = TRUE,
                             explain       = TRUE,
                             unmatched     = "drop",     # or "keep"
                             verbose       = FALSE)
nrow(prep$variants)
```

`unmatched` controls what happens to variants outside every scheme's scope:

- `"drop"` — remove them (default when `apply = TRUE`).
- `"keep"` — leave them in `prep$variants` as-is.

Multiple filter runs stack in `prep$filter_explanation`, keyed by the concatenated scheme names:

```{r}
names(prep$filter_explanation)
```

# 4. Scheme grammar

## 4.1 `rule()` — atomic predicate

Simplest form: match a column against values.

```{r}
rule(Gene = c("TP53","BRCA1"))                       # match any of these
rule(Somatic_status = "somatic")                     # single value
```

Predicate-function form (`rule(VAF_percentage = function(x) x > 5)`) is supported by the evaluator but doesn't have a printable label; use it for numeric-range gates and pair with an explicit `label = "VAF > 5"`.

## 4.2 `allOf()` / `anyOf()` / `not()` — combinators

```{r}
allOf(rule(Gene = "TP53"), rule_somatic(), rule_truncating())      # AND
anyOf(rule_frameshift(), rule_nonsense(), rule_splice_donor())     # OR
not(rule_ch())                                                     # negation
```

`allOf` is aliased as `alt_all`, `anyOf` as `alt_any`, and there's a `alt_any_n(rules, n)` for "at least n of the rules pass".

## 4.3 `ctdna_create_scheme()` — quick constructor

For "gene list + criteria" schemes with no negation or branches:

```{r}
q <- ctdna_create_scheme(
  name     = "quick_TSG",
  genes    = c("TP53","RB1","PTEN"),
  criteria = c("truncating","missense"),  # from: truncating, missense, clinvar_path, cnv_del, cnv_amp, lgr, fusion, somatic, germline, rare
  combine  = "any")
visualized_scheme(q)
```

## 4.4 `create_filtering_scheme()` — full grammar

For everything else: negation, nested combinators, named branches, scope + gate + branch structure.

```{r, eval = FALSE}
create_filtering_scheme(
  # scope
  rule(Gene = c("TP53","RB1","PTEN")),

  # gate (unconditional exclusion)
  not(rule_ch()), not(rule_clinvar_benign()),

  # branches (named alternatives; PASS if any branch matches)
  branches = list(
    germline_PLP = allOf(rule_germline(), rule_clinvar_path()),
    somatic_LoF  = allOf(rule_somatic(),  rule_truncating())),

  name        = "custom_TSG",
  description = "Custom TSG scheme",
  frozen      = FALSE,        # TRUE = schema-locked, can't be modified
  overwrite   = TRUE)
```

## 4.5 `visualized_scheme()` — inspect

Prints a tree of scope / gate / branches:

```{r}
visualized_scheme(TSG_Tier1())
```

## 4.6 `ctdna_test_scheme()` — dry-run against a prep

Reports how many variants pass at each level, without modifying the prep.

```{r}
head(ctdna_test_scheme("TSG_Tier2", df = prep$variants))
```

# 5. Scheme catalog

Schemes are persisted in an in-session catalog (and optionally a user-config file). Full CRUD:

```{r}
ctdna_filter_schemes_list()
```

- `ctdna_filter_scheme_get(name)` — retrieve one.
- `ctdna_filter_scheme_show(name)` — pretty-print one.
- `ctdna_filter_scheme_save(name, scheme, path = NULL)` — persist to disk.
- `ctdna_filter_scheme_load(name, path = NULL)` — read from disk.
- `ctdna_filter_scheme_delete(name)` — remove from catalog.
- `ctdna_filter_reset_general()` — reset the catalog to built-ins.

# 6. Landscape plots

## 6.1 `ctdna_oncoprint()`

Gene × patient heatmap with alteration glyphs.

```{r, fig.height=6}
op <- ctdna_oncoprint(
  prep,
  gene_sets     = list(TSG   = c("TP53","RB1","PTEN"),
                       HRR14 = ctdna_gene_set("HRR14")),
  visit         = "baseline",     # or a specific visit label
  filter_scheme = NULL,           # NULL = no scheme filtering
  wrap          = "Indication",   # facet by column name; NULL = no wrap
  engine        = "auto",         # "ComplexHeatmap" or "ggplot"
  title         = "Baseline landscape")
op$plot
```

Return object: `list(plot, patient_data, matrix, gene_sets, table)`.

## 6.2 `ctdna_alteration_grid()`

Gene × outcome-group prevalence matrix.

```{r, fig.height=4}
gr <- ctdna_alteration_grid(
  prep,
  gene_sets = list(TSG   = c("TP53","RB1","PTEN"),
                   HRR14 = ctdna_gene_set("HRR14")),
  scheme    = "three",          # "two" = R/NR; "three" = R/NR/undefined; "four" = adds SD
  wrap      = "Indication",
  title     = "Prevalence by response")
gr$plot
```

# 7. Trajectory plots

All work off `prep$samples` (per-visit) or `prep$assessments` (tumor size). Each function returns a ggplot object directly, except `ctdna_plot_longitudinal()` which returns a small list with `$plot` and a `$stats` frame.

## 7.1 `ctdna_plot_baseline()`

TF at baseline, grouped by dose. Handles `group_by`, `subgroup_by`, per-group Wilcoxon or Kruskal p-values, log-scale y, etc.

```{r}
ctdna_plot_baseline(prep)
```

## 7.2 `ctdna_plot_reduction()`

TF change from baseline vs RECIST × dose.

```{r}
ctdna_plot_reduction(prep)
```

## 7.3 `ctdna_plot_longitudinal()`

Subject-level TF trajectories across visits.

```{r}
lon <- ctdna_plot_longitudinal(prep)
lon$plot
```

## 7.4 `ctdna_plot_pct_change_by_dose_time()`

%ΔTF grouped by dose × visit. Note: takes a per-sample **data.frame** as first argument, not `prep`.

```{r, eval = FALSE}
ctdna_plot_pct_change_by_dose_time(prep$samples)
```

## 7.5 `ctdna_plot_scatter()`

Generic scatter with correlation stats. `x` / `y` can be raw column names or opts keys (e.g. `"maxvaf"` resolves to whatever column is set in `ctdna_opts("maxvaf")`).

```{r}
ctdna_plot_scatter(prep, x = "maxvaf", y = "methyl")
```

## 7.6 `ctdna_boxplot()`

Generic boxplot with per-group p-values. `y` must be a numeric column of the target frame, `x` a categorical one.

```{r}
ctdna_boxplot(prep, y = "cfDNA_ng", x = "RECIST")
```

To filter/reorder x-axis categories, preprocess `prep$samples` (or whichever slot) with `filter()` + `factor(levels =)` before plotting — not `scale_x_discrete()` post-hoc.

# 8. Cross-modality plots

Each takes **two frames** (ctdna metrics + orthogonal platform data), not `prep`. Extract the two frames you need before plotting.

## 8.1 `ctdna_plot_vs_tmb()`

TF vs TMB. Both live in `prep$samples`.

```{r, eval = FALSE}
ctdna_plot_vs_tmb(prep$samples, prep$samples)
```

## 8.2 `ctdna_plot_vs_mutation()`

TF vs tissue mutation status for a specific gene. Requires a `mut_df` frame with per-patient/gene mutation calls.

```{r, eval = FALSE}
ctdna_plot_vs_mutation(prep$samples, sim$dnaseq, gene = "TP53")
```

## 8.3 `ctdna_plot_vs_ihc()`

TF vs an IHC marker score.

```{r, eval = FALSE}
ctdna_plot_vs_ihc(prep$samples, sim$ihc, marker = "PD-L1")
```

## 8.4 `ctdna_plot_vs_expression()`

TF vs RNAseq expression of a specific gene.

```{r, eval = FALSE}
ctdna_plot_vs_expression(prep$samples, sim$rnaseq$tpm, gene = "TP53")
```

# 9. Statistics

Helpers for computing and formatting p-values, correlation stats, and group tests. Most extract stats from a ggplot object; a few operate on raw vectors / frames.

## 9.1 P-value labels — `ctdna_pval_label(p)`

Given a ggplot returned by `ctdna_plot_baseline()` / `ctdna_plot_reduction()` / `ctdna_boxplot()`, extract the annotated p-values.

```{r, eval = FALSE}
p <- ctdna_plot_baseline(prep)
ctdna_pval_label(p)
```

## 9.2 `scatter_stats(p)`

Extract Pearson / Spearman / Kendall stats from a scatter plot.

```{r}
s <- ctdna_plot_scatter(prep, x = "maxvaf", y = "methyl")
scatter_stats(s)
```

## 9.3 `longitudinal_stats(p)`

Extract per-subject slope stats from a longitudinal plot.

```{r, eval = FALSE}
l <- ctdna_plot_longitudinal(prep)
longitudinal_stats(l)
```

## 9.4 Pairwise Wilcoxon — `ctdna_pairwise_ranksum(data, value, group, method)`

```{r}
head(ctdna_pairwise_ranksum(prep$samples,
                             value  = "cfDNA_ng",
                             group  = "RECIST",
                             method = "wilcox"))
```

## 9.5 Paired test — `ctdna_paired_test(...)`

Paired Wilcoxon or paired t-test. Signature depends on version; see `?ctdna_paired_test`.

## 9.6 Correlation test — `ctdna_cor_test(...)`

Wrapped `cor.test()` with a tidy row output.

## 9.7 Concordance / agreement

- `ctdna_concordance(x, y)` — top-level agreement between two callers.
- `ctdna_concordance_test(x, y)` — inferential version.
- `ctdna_cohen_kappa(x, y)`, `ctdna_weighted_kappa(x, y)` — categorical agreement.
- `concordance_stats(x, y)` — low-level stat calculator.

## 9.8 `ctdna_metric_at(pred, truth, metric, at)`

Evaluate a metric at a threshold.

## 9.9 Summary and table

```{r, eval = FALSE}
ctdna_summary(prep)      # one-line rowcount summary
```

`ctdna_table(prep, slot, ...)` — quick frequency tabulation with sensible defaults.

## 9.10 `ctdna_run_group_test()`

Runs a per-group test and returns a tidy frame — used internally by plot functions.

# 10. Themes, colors, and scales

## 10.1 Theme

`ctdna_theme(base_size = 12, ...)` — base theme applied by every ctdnaTM plot. Add to external ggplots to match style.

```{r, eval = FALSE}
library(ggplot2)
ggplot(mtcars, aes(mpg, hp)) + geom_point() + ctdna_theme()
```

## 10.2 Color palettes

`ctdna_colors()` returns the full palette registry (a large list of named palettes). `ctdna_dose_palette(levels)` builds a palette for an arbitrary dose vector.

```{r, eval = FALSE}
ctdna_colors()                             # inspect the palette registry
ctdna_dose_palette(c("6 mg/kg", "8 mg/kg"))
```

## 10.3 ggplot scales

- `ctdna_scale_dose()` / `ctdna_scale_dose_fill()` / `ctdna_scale_dose_shape()` — for a `Dose` aesthetic.
- `ctdna_scale_recist()` / `ctdna_scale_recist_fill()` — for a `RECIST` aesthetic.

Use just like `scale_color_manual()`:

```{r, eval = FALSE}
ggplot(prep$samples, aes(Visit_num, tumor_fraction, color = Dose)) +
  geom_line() +
  ctdna_scale_dose()
```

## 10.4 `ctdna_stratify_recist()`

Collapse the 4-level RECIST column into 2-, 3-, or 4-level buckets (R/NR, R/NR/undefined, or CR-PR/SD/PD/undefined).

```{r}
ctdna_stratify_recist(prep$clinical$RECIST, recist_grouping = "three") |> table()
```

# 11. Utility functions

## 11.1 `ctdna_compute_mr()`

Molecular response classification per patient.

## 11.2 `ctdna_floor_tf()`

Impute below-detection TF values to a floor (default `1e-3`).

## 11.3 `ctdna_gene_set()`

Return a named gene set. Built-in sets: `HRR14`, `TSG`, `Cell_Cycle`, `MMR`, `PI3K`, `RTK`, `TP53_pathway`.

```{r}
ctdna_gene_set("HRR14")
```

## 11.4 `ctdna_alteration_types()`

The canonical alteration-type vocabulary used by the oncoprint glyphs.

```{r}
ctdna_alteration_types()
```

## 11.5 `ctdna_ratio()`

Compute a ratio with graceful `NA` handling.

## 11.6 `ctdna_coerce()`

Safe type coercion with `NA` propagation on failure.

## 11.7 `ctdna_to_log2tpm()`

Log2(TPM + 1) transformation for RNAseq expression.

# 12. Multi-modal helpers

## 12.1 `ctdna_assemble_modalities()`

Combine several data-frames representing different modalities into a single long/wide frame keyed by `Patient_ID` and `Visit_name`.

## 12.2 `ctdna_merge_modalities()`

Join a modality frame to `prep$samples` on sample/patient/visit keys.

## 12.3 Population frequency annotation

- `ctdna_annotate_population_freq(prep)` — annotate `prep$variants` with population frequencies (gnomAD/1000G-style).
- `ctdna_load_popfreq_cache(path)` / `ctdna_save_popfreq_cache(path)` — cache the pop-freq lookup between runs.

# 13. Pipeline

## 13.1 `ctdna_pipeline()`

Runs every registered deliverable.

```{r, eval = FALSE}
res <- ctdna_pipeline(prep,
                      scheme        = "three",
                      visit         = "baseline",
                      wrap          = "Indication",
                      gene_sets     = list(TSG = c("TP53","RB1","PTEN"),
                                           HRR14 = ctdna_gene_set("HRR14")),
                      filter_scheme = c("TSG_Tier1","HRR14"),
                      engine        = "auto",
                      opts          = list(oncoprint = list(title = "Custom title"),
                                           vs_ihc    = list(skip = TRUE)),
                      verbose       = FALSE)
```

## 13.2 `ctdna_pipeline_steps()`

List every registered deliverable (11 total).

```{r}
ctdna_pipeline_steps()
```

## 13.3 `ctdna_run_pipeline()`

Lower-level pipeline runner used internally by `ctdna_pipeline()`; can register custom pipeline steps.

# 14. Configuration

## 14.1 `ctdna_opts()`

Read or set package-wide options.

```{r}
# ctdna_opts() with no args prints the full opts registry
ctdna_opts("baseline")
```

## 14.2 `ctdna_opts_choices()`

List the allowed values for enumerated options.

## 14.3 `ctdna_opts_save()` / `ctdna_opts_load()`

Persist / restore your option set (e.g. across R sessions).

## 14.4 `ctdna_filter_opts()`

Filter-engine options.

# 15. Reference: predicate library

Predicates organized by domain. Every one is a `ctdna_rule` object usable inside `allOf`/`anyOf`/`not`/`create_filtering_scheme`.

## 15.1 Variant type

`rule_is_SNV()`, `rule_is_Indel()`, `rule_is_CNV()`, `rule_is_Fusion()`, `rule_is_LGR()`, `rule_is_LGR_deleterious()`.

## 15.2 Consequence

`rule_missense()`, `rule_nonsense()`, `rule_frameshift()`, `rule_truncating()`, `rule_synonymous()`, `rule_inframe_indel()`, `rule_start_lost()`, `rule_stop_lost()`, `rule_promoter()`, `rule_splice_donor()`, `rule_splice_acceptor()`, `rule_splice_region()`, `rule_splice_event_mc()`.

## 15.3 Somatic / germline / CH

`rule_somatic()`, `rule_germline()`, `rule_ch()`.

## 15.4 ClinVar / functional

`rule_clinvar_path()`, `rule_clinvar_benign()`, `rule_clinvar_vus()`, `rule_clinvar_conflict()`, `rule_deleterious()`.

## 15.5 CNV / SV

`rule_cnv_amp()`, `rule_cnv_focal_amp()`, `rule_cnv_aneuploid_amp()`, `rule_cnv_loh_del()`, `rule_cnv_homozyg_del()`, `rule_amp_thresh(threshold)`, `rule_biallelic()`, `rule_lgr_deletion()`, `rule_lgr_inversion()`, `rule_lgr_tandem_duplication()`.

## 15.6 Sample-level

`rule_sample_success()`, `rule_sample_fail()`.

## 15.7 Population frequency

`rule_common_001()`, `rule_rare_001()`, `rule_rare_0001()`.

## 15.8 Special

`rule_reversion()`, `rule_BRCA2_K3326()`.

## 15.9 Gene set

`rule_genes_TSG()`, `rule_genes_HRR14()`, `rule_genes_MMR()`, `rule_genes_PI3K()`, `rule_genes_RTK()`, `rule_genes_Cell_Cycle()`, `rule_genes_TP53_pathway()`.

## 15.10 Indication-specific

`rule_ind_BRCA()`, `rule_ind_CRC()`, `rule_ind_GBM()`, `rule_ind_HNSCC()`, `rule_ind_NSCLC()`, `rule_ind_SCLC()`, `rule_ind_mCRPC()`.

# 16. Package meta

## 16.1 `ctdna_help()`

Open the help index.

## 16.2 `ctdna_logo()`

Path to the shipped hex logo.

```{r}
ctdna_logo()
```

Use in Rmd:

```{r, eval = FALSE}
knitr::include_graphics(ctdna_logo())
```

## 16.3 `ctdna_methods()`

Enumerate every registered analysis method (used by the pipeline).

## 16.4 `ctdna_register_method()`

Register a custom method into the pipeline registry.

## 16.5 `ctdna_check_for_updates()`

Check the CRAN / GitHub release for a newer version.

## 16.6 `ctdna_cleanup()`

Clear session-scoped ctdnaTM caches.

# Session info

```{r}
sessionInfo()
```
