---
title: "ctdnaTM — Filter system walkthrough (v0.75.10)"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

```{r logo, echo = FALSE, results = "asis"}
logo_path <- system.file("extdata", "logo.png", package = "ctdnaTM")
if (nzchar(logo_path) && file.exists(logo_path))
  cat(sprintf('<img src="%s" style="float:right; width:120px; margin-left:1em" alt="ctdnaTM logo"/>\n\n', logo_path))
```

```{r setup, include = FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
if (!requireNamespace("ctdnaTM", quietly = TRUE))
  install.packages(
    "/mnt/code/Hn/lib/ctdnaTM_0.43.8.tar.gz",
    repos = NULL, type = "source")
library(ctdnaTM)
```

# About this document

Walkthrough of the **ctdnaTM filter system**. The filter system itself
is unchanged in v0.75.10 (no behavioural changes to rules, schemes, or
`ctdna_filter_apply()`); this Rmd is the same content as the v0.38.0
walkthrough with the install URL updated.

```{r data_setup}
sim  <- ctdna_make_mock_study(n_patients = 60, seed = 22)
prep <- ctdna_prepare(sim, verbose = FALSE)
gen  <- prep$genomic
cat("Genomic rows:", nrow(gen), "\n")
```

---

# 1. The two stages of `ctdna_filter_apply()`

```r
ctdna_filter_apply(df,
                   filter_scheme    = scheme_basic(),    # NULL = no variant filter
                   sample_qc_enable = NULL)              # NULL = use opts default
```

Two independent stages run in this order:

| Stage | Controlled by | Reads from |
|---|---|---|
| **Stage 0** — Sample QC | `sample_qc_enable` arg (or `ctdna_opts("sample_qc_filter")` fallback) | `Sample_status`, cfDNA / plasma volume minimums |
| **Stage 1** — Variant filter | `filter_scheme` arg | The full rule language on each alteration row |

## Three useful modes

```{r}
# Mode 1: QC only (no variant filtering)
df_qc <- ctdna_filter_apply(gen, filter_scheme = NULL,
                              sample_qc_enable = TRUE)
nrow(df_qc)

# Mode 2: Variant filter only (skip QC)
df_var <- ctdna_filter_apply(gen, filter_scheme = scheme_basic(),
                               sample_qc_enable = FALSE)
nrow(df_var)

# Mode 3 (default): both stages
df_both <- ctdna_filter_apply(gen, filter_scheme = scheme_basic())
nrow(df_both)
```

The returned frame carries `attr(., "ctdna_filter_used")` documenting
which mode ran.

```{r}
attr(df_both, "ctdna_filter_used")
```

---

# 2. The rule language

A rule is a function that takes a data frame and returns a logical
vector. The library ships ~80 named rules organised by category.

## Building rules

```{r}
# A column-equality rule
my_rule <- rule(column = "Gene", op = "==", value = "TP53")

# A built-in predicate
rule_clinvar_path()

# Compose with combinators
allOf(rule_clinvar_path(), rule_truncating())
anyOf(rule_clinvar_path(), rule_clinvar_vus())
`not`(rule_synonymous())
```

## Evaluating a rule

The bare-bones path:

```{r}
mask <- ctdnaTM:::.eval_rule(rule_clinvar_path(), gen)
sum(mask)
```

The user-facing path:

```{r}
df_path <- ctdna_filter_apply(gen, filter_scheme = rule_clinvar_path(),
                                 sample_qc_enable = FALSE)
nrow(df_path)
```

---

# 3. Rules by category

| Category | Examples |
|---|---|
| **Variant type** | `rule_is_SNV()`, `rule_is_Indel()`, `rule_is_CNV()`, `rule_is_Fusion()`, `rule_is_LGR()` |
| **Consequence** | `rule_missense()`, `rule_nonsense()`, `rule_frameshift()`, `rule_synonymous()`, `rule_splice_donor()`, `rule_splice_acceptor()`, `rule_splice_region()`, `rule_splice_event_mc()`, `rule_inframe_indel()`, `rule_start_lost()`, `rule_stop_lost()`, `rule_promoter()`, `rule_truncating()` |
| **ClinVar** | `rule_clinvar_path()`, `rule_clinvar_benign()`, `rule_clinvar_vus()`, `rule_clinvar_conflict()` |
| **Origin** | `rule_germline()`, `rule_somatic()`, `rule_ch()` (clonal haematopoiesis) |
| **Frequency** | `rule_common_001()`, `rule_rare_001()`, `rule_rare_0001()` |
| **Gene sets** | `rule_genes_HRR14()`, `rule_genes_TSG()`, `rule_genes_PI3K()`, `rule_genes_RTK()`, `rule_genes_MMR()`, `rule_genes_Cell_Cycle()`, `rule_genes_TP53_pathway()` |
| **Indication** | `rule_ind_BRCA()`, `rule_ind_NSCLC()`, `rule_ind_SCLC()`, `rule_ind_CRC()`, `rule_ind_HNSCC()`, `rule_ind_GBM()`, `rule_ind_mCRPC()` |
| **CNV specifics** | `rule_cnv_amp()`, `rule_cnv_focal_amp()`, `rule_cnv_aneuploid_amp()`, `rule_cnv_homozyg_del()`, `rule_cnv_loh_del()`, `rule_amp_thresh()`, `rule_biallelic()` |
| **LGR specifics** | `rule_lgr_deletion()`, `rule_lgr_inversion()`, `rule_lgr_tandem_duplication()`, `rule_is_LGR_deleterious()` |
| **Sample** | `rule_sample_success()`, `rule_sample_fail()` |
| **Special** | `rule_deleterious()`, `rule_reversion()`, `rule_BRCA2_K3326()` |

---

# 4. Pre-built schemes

## `scheme_basic()`

Cancer-actionable alterations: somatic + impactful + filter on
suspected germline / CH / synonymous.

```{r}
df_basic <- ctdna_filter_apply(gen, filter_scheme = scheme_basic(),
                                 sample_qc_enable = FALSE)
nrow(df_basic)
```

## `scheme_HRR()`

HRR-gene focus: HRR14 panel + ClinVar-pathogenic / truncating /
LGR-deleterious / known reversions.

```{r}
df_hrr <- ctdna_filter_apply(gen, filter_scheme = scheme_HRR(),
                               sample_qc_enable = FALSE)
nrow(df_hrr)
```

## `scheme_TSG()`

Tumor-suppressor focus.

```{r}
df_tsg <- ctdna_filter_apply(gen, filter_scheme = scheme_TSG(),
                               sample_qc_enable = FALSE)
nrow(df_tsg)
```

---

# 5. Visualizing a scheme

```{r eval = FALSE}
visualized_scheme(scheme_basic())     # ASCII tree
ctdna_filter_scheme_show(scheme_HRR()) # detailed printout
```

---

# 6. Scheme management

| Function | Purpose |
|---|---|
| `ctdna_filter_schemes_list()` | List saved schemes |
| `ctdna_filter_scheme_get()` | Retrieve a saved scheme |
| `ctdna_filter_scheme_save()` | Persist a scheme to disk |
| `ctdna_filter_scheme_load()` | Load a scheme from disk |
| `ctdna_filter_scheme_delete()` | Remove a saved scheme |
| `create_filtering_scheme()` | Builder helper |
| `ctdna_test_scheme()` | Run sanity tests on a scheme |

---

# 7. Filter options

`ctdna_filter_opts()` exposes the gnomAD-rule defaults, the default
sample-QC behaviour, and other knobs.

```{r}
ctdna_filter_opts()
```

Reset to package defaults:

```{r eval = FALSE}
ctdna_filter_reset_general()
```

---

# 8. The visit-filtering question (important)

`scheme_basic()` and the other variant filters do NOT filter by
`Visit_name`. The "alterations only at C1D1" behaviour comes from the
upstream assembly step `.gh_build_genomic()` in `R/platforms.R`,
which is by design for GH Infinity (variant calls happen once at
baseline; on-treatment visits use methylation / TF tracking).

To analyse longitudinal data, look at `prep$ctdna` (longitudinal
TF / methylation), not `prep$genomic` (baseline-only variant calls).

---

# Session info

```{r session, echo = FALSE}
sessionInfo()
```
