---
title: "ctdnaTM: end-to-end walkthrough"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{ctdnaTM: end-to-end walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5, dpi = 100,
                      message = FALSE, warning = FALSE)
```

```{r logo, echo=FALSE, out.width='180px', fig.align='left'}
p <- tryCatch(ctdnaTM::ctdna_logo(), error = function(e) "")
if (nzchar(p) && file.exists(p)) knitr::include_graphics(p)
```

A single linear pass through `ctdnaTM`: **prepare → QC → filter → plots → statistics → pipeline**. For every function argument-by-argument, see `vignette("ctdnaTM_function_tour")`. For how the filter engine works internally, see `vignette("ctdnaTM_filter_internals")`.

# 1. Install & load

```{r}
library(ctdnaTM)
packageVersion("ctdnaTM")
```

The package is built around **one central object** — a `ctdna_prep` list called `prep` by convention. Every downstream function (`ctdna_sample_qc`, `ctdna_variant_filter`, every plot, the pipeline) takes `prep` as its first argument and does not need direct access to the raw data files. This makes the analysis reproducible: two analysts running the same `ctdna_prepare()` call get the same `prep`, and the same plots.

# 2. Assemble the prep object

`ctdna_prepare()` reads a Guardant Health **Infinity Report** (a wide, one-row-per-variant CSV/TSV) together with any **ADaM frames** you pass (adsl / adrs / adtr) and returns a named list with these slots:

- `$clinical` — one row per patient, union of every ADaM frame you provided.
- `$samples` — one row per Sample_ID (per-visit metadata: QC status, cfDNA ng, plasma volume, TMB, TF, detection status, …).
- `$variants` — one row per called alteration from the Infinity report, verbatim. Not filtered at prep time.
- `$assessments` — long-format tumor size assessments from ADTR (used by `ctdna_boxplot`, `ctdna_plot_scatter`, and the tumor-size plots).
- `$adrs` — response data if you passed an ADRS frame.
- `$dictionary` — a stable snapshot of every column name found in every frame; used internally by helpers such as `.read_per_patient()`.
- `$qc_removed` — populated by `ctdna_sample_qc()` after you run it; empty until then.

Use `ctdna_make_mock_study()` to generate a full mock dataset so this vignette is self-contained:

```{r}
sim <- ctdna_make_mock_study(n_patients = 60, seed = 1)
# oncoprint's wrap = "Indication" needs a lower-case `indication` column;
# the mock ships with Cancertype so we copy it over.
sim$clinical$indication <- sim$clinical$Cancertype

prep <- ctdna_prepare(
  infinity_report = sim$infinity_report,                             # variant + sample layer
  adam            = list(adsl = sim$clinical, adtr = sim$adtr),      # union across all frames
  verbose         = FALSE)

class(prep)
setdiff(names(prep), "dictionary")
```

**Key facts about `ctdna_prepare()`:**

- **`prep$variants` is the Infinity report verbatim** (plus a `Baseline` boolean column). No rows are filtered at prep time. If a row is in your CSV, it is in `prep$variants`.
- **`prep$clinical` is the union of every ADaM frame's patient IDs.** A patient present only in adrs or only in adtr still gets a row in `prep$clinical`, with columns coalesced across frames in list order (first non-NA wins).
- **`ctdna_prepare()` no longer runs sample QC automatically.** You call `ctdna_sample_qc()` yourself, when you want, and only what fails your defined criteria is removed.

# 3. Sample QC

`ctdna_sample_qc()` removes samples that fail the QC criteria you set, cascades the removal to any variants tied to those samples, and archives the removed rows in `prep$qc_removed`. It is idempotent — calling it twice on the same prep is a no-op the second time.

```{r}
prep <- ctdna_sample_qc(prep, verbose = FALSE)

# Row counts after QC
nrow(prep$samples)
nrow(prep$variants)
```

Default criteria come from `ctdna_opts()`:

- `sample_status` — a whitelist of acceptable `Sample_status` values. Default: `SUCCESS`, `Success`, `success`, `PASS`, `Pass`, `pass`. Anything else fails.
- `plasma_min` — minimum `Plasma_ml_input`. Default `0` (off).
- `cfdna_min` — minimum `cfDNA_ng`. Default `0` (off).

Pass explicit values to override:

```{r, eval = FALSE}
prep <- ctdna_sample_qc(prep,
                        sample_status = c("SUCCESS"),
                        plasma_min    = 2.0,
                        cfdna_min     = 5.0)
```

**Only variants tied to a failed sample are cascaded out.** Everyone else stays. `prep$clinical` is never touched by sample QC — a patient whose only sample failed still appears in `$clinical` (they'll show up as all-wildtype in the oncoprint).

# 4. Variant filtering

The single entry point is `ctdna_variant_filter(prep, filter_scheme, apply, explain)`. Both `apply` and `explain` are mandatory — no defaults, so you decide up-front what you want.

| `apply` | `explain` | effect |
|:-:|:-:|:--|
| `TRUE`  | `TRUE`  | `prep$variants` is filtered in-place; an annotated snapshot lives at `prep$filter_explanation$<sig>` |
| `TRUE`  | `FALSE` | filtered in-place, no snapshot (fast) |
| `FALSE` | `TRUE`  | `prep$variants` unchanged; snapshot only (dry-run) |
| `FALSE` | `FALSE` | error — nothing to do |

The `<sig>` key is the concatenation of scheme names, so multiple filter runs coexist under distinct keys.

```{r}
prep <- ctdna_variant_filter(
  prep,
  filter_scheme = c("TSG_Tier2", "HRR14"),
  apply         = TRUE,
  explain       = TRUE,
  verbose       = FALSE)

# Filtered variants
nrow(prep$variants)

# Snapshot key + a peek at the annotated rows
sig <- names(prep$filter_explanation)[1]
sig
head(sort(table(prep$filter_explanation[[sig]]$filtering_criteria,
                useNA = "ifany"), decreasing = TRUE), 6)
```

The `filtering_criteria` column carries a **branch label** telling you *why* each row passed or failed:

- `PASS::TSG_Tier2(somatic_LoF)` — passed via the somatic loss-of-function branch of Tier 2.
- `PASS::TSG_Tier2(somatic_LoF), HRR14(deleterious_LGR)` — passed both schemes; branch labels tell you which rule fired in each.
- `FAIL::TSG_Tier2, HRR14` — variant was in both schemes' scope but did not pass either.
- `NA` — variant was in **no** scheme's scope (e.g. an off-panel gene). If `apply = TRUE` with default `unmatched = "drop"`, these rows are dropped from `prep$variants`.

# 5. Built-in schemes

Four biomarker schemes come with the package. All of them drop ClinVar-benign and CHIP (clonal hematopoiesis) variants unconditionally — those never make it through the scope gate.

- **`TSG_Tier1`** — Tumor suppressor genes; most permissive. Passes any *impactful* somatic alteration in TP53 / RB1 / PTEN, plus germline pathogenic / likely-pathogenic, plus deleterious LGRs.
- **`TSG_Tier2`** — Same scope as Tier 1 but stricter: requires the somatic alteration to be a **loss-of-function** event (truncating / frameshift / splice-donor / splice-acceptor / start-lost / stop-lost).
- **`TSG_Tier3`** — Most restrictive of the three. TP53 / RB1 / PTEN only; germline P/LP; somatic LoF; somatic ClinVar-P/LP missense / inframe; LOH or homozygous del. No LGR / fusion / rare-synonymous branches.
- **`HRR14`** — 14 canonical homologous-recombination-repair genes (BRCA1/2, ATM, PALB2, CHEK2, etc.). Same three branches as Tier 2 plus an intra-genic LGR branch.
- **`scheme_basic`** — Everything in the panel with any impactful (nonsyn / LGR / CNV) alteration. Broad screening.

Inspect any scheme:

```{r}
visualized_scheme(TSG_Tier2())
```

# 6. Build your own scheme

Two ways, low-level and high-level.

**High-level (`ctdna_create_scheme`)** — for simple "gene list + criteria" schemes:

```{r}
myscheme <- ctdna_create_scheme(
  name     = "my_TSG_lite",
  genes    = c("TP53", "PTEN"),
  criteria = c("truncating", "missense"),  # any of: truncating, missense, clinvar_path, cnv_del, cnv_amp, lgr, fusion, somatic, germline, rare
  combine  = "any")
visualized_scheme(myscheme)
```

**Low-level (`create_filtering_scheme`)** — for the full grammar with negation, nesting, and named branches:

```{r}
my_tsg <- create_filtering_scheme(
  # scope: at least one gene in {TP53, RB1, PTEN}
  rule(Gene = c("TP53","RB1","PTEN")),

  # gate: never CHIP, never ClinVar-benign
  not(rule_ch()),
  not(rule_clinvar_benign()),

  branches = list(
    germline_PLP    = allOf(rule_germline(), rule_clinvar_path()),
    somatic_LoF     = allOf(rule_somatic(),  rule_truncating()),
    LOH_homozyg_del = allOf(rule_somatic(),
                            anyOf(rule_cnv_loh_del(),
                                  rule_cnv_homozyg_del()))),

  name        = "my_tsg",
  description = "TP53/RB1/PTEN, three named branches, no CHIP/benign.",
  overwrite   = TRUE)

visualized_scheme(my_tsg)
```

Test the scheme against your prep before committing:

```{r}
ctdna_test_scheme("my_tsg", df = prep$variants) |> head()
```

# 7. Landscape plots

Two functions build gene × patient matrices for the two flagship deliverables.

**`ctdna_oncoprint(prep, gene_sets, ...)`** — one column per patient, one row per gene, one glyph per alteration. Handles multi-hit patients (draws stacked glyphs), and can wrap by indication or other clinical covariates.

```{r, fig.height=6}
op <- ctdna_oncoprint(
  prep,
  gene_sets     = list(TSG   = c("TP53","RB1","PTEN"),
                       HRR14 = ctdna_gene_set("HRR14")),
  visit         = "baseline",
  filter_scheme = NULL,                         # no scheme gate here
  wrap          = "Indication",
  engine        = "auto",
  title         = "Baseline landscape")
op$plot
```

The wider cohort (v0.75.8+) means every patient in `prep$clinical` with an Indication contributes a column, even if they have no altered gene in the panel — they show as an all-wildtype stripe. Set `wrap = NULL` to disable the panel split.

**`ctdna_alteration_grid(prep, gene_sets, ...)`** — one row per gene, columns split by an outcome (defaults to RECIST-based responder / non-responder). Cells show alteration prevalence per group. Useful for "is this gene enriched in responders?" questions.

```{r, fig.height=4}
gr <- ctdna_alteration_grid(
  prep,
  gene_sets = list(TSG   = c("TP53","RB1","PTEN"),
                   HRR14 = ctdna_gene_set("HRR14")),
  scheme    = "three",                          # 3-column: R / NR / undefined
  title     = "Alteration prevalence by response")
gr$plot
```

# 8. Trajectory plots

Six plot functions work off `prep$samples` (per-visit metrics) or `prep$assessments` (tumor size long-format). Each returns a ggplot object directly (except `ctdna_plot_longitudinal()`, which returns a small list with `$plot` and a stats frame).

| Function | X axis | Y axis | Data source |
|---|---|---|---|
| `ctdna_plot_baseline()`    | Dose / group | TF at baseline (log10)                    | `prep$samples`     |
| `ctdna_plot_reduction()`   | Dose × RECIST | TF-change vs baseline                    | `prep$samples`     |
| `ctdna_plot_longitudinal()`| Visit         | TF (log10) with subject lines            | `prep$samples`     |
| `ctdna_plot_pct_change_by_dose_time()` | Dose × Visit | %ΔTF                              | any per-sample df  |
| `ctdna_plot_scatter()`     | any numeric   | any numeric                              | any prep frame     |
| `ctdna_boxplot()`          | any categorical | any numeric                            | any prep frame     |

`ctdna_boxplot()` and `ctdna_plot_scatter()` are general — pass any prep-slot numeric/categorical pair and they build a boxplot or scatter, including per-group p-values.

```{r}
ctdna_plot_baseline(prep)
```

To filter or reorder x-axis categories in `ctdna_boxplot()`, preprocess the frame before plotting rather than using `scale_x_discrete()` post-hoc — otherwise the significance-bracket layer stays in coordinate space and leaves phantom p-values.

```{r, eval=FALSE}
keep <- c("Cycle 1 Day 1", "Cycle 2 Day 1", "Cycle 3 Day 1")
prep_sub <- prep
prep_sub$assessments <- prep_sub$assessments |>
  dplyr::filter(Tumor_visit_name %in% keep) |>
  dplyr::mutate(Tumor_visit_name = factor(Tumor_visit_name, levels = keep))
ctdna_boxplot(prep_sub, y = "tumor_size_change", x = "Tumor_visit_name")
```

# 9. Cross-modality plots

Four `vs_*` plots correlate ctDNA with another platform. All work off `prep$samples` plus a second frame you attach with `ctdna_prep_add()` (or that comes from the ADaM ingestion).

- `ctdna_plot_vs_tmb()` — TF vs TMB (both already in `$samples`; no extra frame needed).
- `ctdna_plot_vs_mutation()` — cfDNA VAF vs tissue VAF (needs `$variants` + a tissue DNAseq frame).
- `ctdna_plot_vs_ihc()` — TF vs an IHC score (needs an IHC frame).
- `ctdna_plot_vs_expression()` — TF vs an RNAseq expression measure (needs an RNAseq frame).

Attach optional modalities:

```{r, eval=FALSE}
prep <- ctdna_prep_add(prep, ihc         = sim$ihc)
prep <- ctdna_prep_add(prep, expression  = sim$rnaseq)
prep <- ctdna_prep_add(prep, dnaseq      = sim$dnaseq)
```

# 10. Statistics

Companion helpers for adding p-values and correlation stats to plots, or for running group tests programmatically.

```{r, eval = FALSE}
# per-group Wilcoxon p-value to overlay on a boxplot
ctdna_pval_label(prep$samples, y = "tumor_fraction", group = "RECIST")

# correlation stats (r, Spearman rho, n) for a scatter overlay
scatter_stats(prep$samples$Plasma_ml_input, prep$samples$cfDNA_ng)

# pairwise Wilcoxon test between every dose level, BH-adjusted
ctdna_pairwise_ranksum(prep$samples$tumor_fraction, prep$samples$Dose)

# concordance between orthogonal calls
ctdna_concordance(x, y, method = "kappa")
```

# 11. The pipeline

`ctdna_pipeline()` runs every registered deliverable and returns a named list of results, each with `$plot`, `$table`, and any per-deliverable extras. The 11 registered deliverables cover the core ctDNA analyses:

```{r}
ctdna_pipeline_steps()
```

Run everything in one call:

```{r, eval = FALSE}
res <- ctdna_pipeline(
  prep,
  scheme        = "three",
  visit         = "baseline",
  wrap          = "Indication",
  gene_sets     = list(TSG   = c("TP53","RB1","PTEN"),
                       HRR14 = ctdna_gene_set("HRR14")),
  filter_scheme = c("TSG_Tier1","HRR14"),
  engine        = "auto",
  verbose       = FALSE)

names(res)              # 11 deliverables
res$oncoprint$plot      # each is $plot + $table + extras
```

Skip specific deliverables via `opts = list(<name> = list(skip = TRUE))`, or pass per-deliverable options via `opts = list(<name> = list(<arg> = <value>))`.

# 12. Themes, colors, scales

Every plot uses `ctdna_theme()` by default. Add it to any external ggplot to match:

```{r, eval = FALSE}
library(ggplot2)
ggplot(mydata, aes(x, y)) + geom_point() + ctdna_theme()
```

Palette helpers:

- `ctdna_colors("dose")`, `ctdna_colors("recist")`, `ctdna_colors("indication")` — named vectors.
- `ctdna_scale_dose()` / `ctdna_scale_dose_fill()` / `ctdna_scale_dose_shape()` — full ggplot scales for the Dose column.
- `ctdna_scale_recist()` / `ctdna_scale_recist_fill()` — same for RECIST.
- `ctdna_dose_palette()` — build a matching palette for a novel dose set.

# Where next

- **`vignette("ctdnaTM_function_tour")`** — every exported function with a runnable example, organized by category.
- **`vignette("ctdnaTM_filter_internals")`** — the filter engine internals: rule objects, evaluator, engine, semantics, predicate library, and how to extend it.
- `?ctdna_prepare`, `?ctdna_variant_filter`, `?ctdna_oncoprint` — full argument docs.
- `ctdna_help()` — quick help index (opens the help page for each core function).

# Session info

```{r}
sessionInfo()
```
