---
title: "ctdnaTM function tour"
author: "Hamid Nikbakht"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ctdnaTM function tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                     fig.width = 7, fig.height = 4.5, dpi = 100,
                     message = FALSE, warning = FALSE)
```

This vignette walks through every user-facing function in `ctdnaTM`, grouped by workflow stage. Every function is invoked at least once; every argument that has a non-default meaning is exercised. The mock study built by `ctdna_make_mock_study()` is used throughout so the whole document runs end-to-end without external data.

# 1. Set up

```{r}
library(ctdnaTM)
sim  <- ctdna_make_mock_study(n_patients = 60, seed = 3)
```

`ctdna_make_mock_study()` returns a list with the same shape as a real Guardant Health Infinity delivery — `infinity_report` (variants), optional panel frames, and a simple `clinical` ADaM frame.

# 2. Prepare and QC

## 2.1 `ctdna_prepare()`

Assembles a `ctdna_prep` object from an Infinity report plus any ADaM frames. In v0.73+, `ctdna_prepare()` no longer runs sample QC by default — `prep$variants` comes back raw.

```{r}
prep <- ctdna_prepare(
  infinity_report = sim$infinity_report,
  adam            = list(adsl = sim$clinical),
  verbose         = FALSE)

class(prep)
names(prep)
nrow(prep$variants)  # raw variants (pre-QC)
```

## 2.2 `ctdna_sample_qc()`

Runs sample-level QC on `prep$samples`, drops failing samples, and cascades to `prep$variants` (variants of dropped samples are removed). Idempotent — calling it a second time drops nothing new.

```{r}
prep <- ctdna_sample_qc(prep, verbose = FALSE)
nrow(prep$variants)  # after QC
prep <- ctdna_sample_qc(prep, verbose = FALSE)
nrow(prep$variants)  # unchanged on rerun
```

# 3. Variant filtering

## 3.1 Grammar overview

Every filter scheme is built from four combinators plus a library of predicates:

- `rule(Column = value)` — atomic predicate. `value` can be a scalar, vector, or `list(op = "!=", value = ...)` for operators (`==`, `!=`, `<`, `<=`, `>`, `>=`, `%in%`, and `==col` for column-to-column comparisons).
- `allOf(a, b, ...)` — AND.
- `anyOf(a, b, ...)` — OR.
- `not(x)` — NOT.

Predicate helpers such as `rule_somatic()`, `rule_truncating()`, `rule_clinvar_path()` are shortcuts for common atomic rules.

```{r}
# Atomic + AND + OR + NOT
r <- allOf(
  rule(Gene = c("TP53","KRAS")),
  not(rule_synonymous()),
  anyOf(rule_somatic(), rule_germline()))
class(r)
```

## 3.2 Built-in schemes

`ctdnaTM` ships four biomarker schemes plus a permissive fallback, keyed by name:

- `TSG_Tier1()` — most permissive TSG (TP53/RB1/PTEN). Any impactful somatic short variant, fusion, or deleterious LGR.
- `TSG_Tier2()` — intermediate TSG (TP53/RB1/PTEN). Somatic LoF, ClinVar-P/LP missense/inframe, LOH or homozygous deletion, or somatic deleterious intra-chromosomal LGR.
- `TSG_Tier3()` — most restrictive TSG (TP53/RB1/PTEN). Germline P/LP, somatic LoF, somatic ClinVar-P/LP + missense/inframe, LOH or homozygous deletion. No LGR, no fusion, no rare-P/LP-synonymous carve-in.
- `HRR14()` — HRR biomarker on 14 HRR genes. ClinVar P/LP, truncating + rare + not-BRCA2-K3326X, homozygous deletion, or somatic deleterious intra-genic LGR.
- `scheme_basic()` — GH Infinity "impactful alterations" ruleset applied to all genes.

All four biomarker schemes drop ClinVar benign and CHIP unconditionally. Each scheme's body is broken into **named branches** so the `filtering_criteria` column tells you which branch fired (e.g. `PASS::TSG_Tier1(somatic_nonsyn)` rather than just `PASS::TSG_Tier1`).

```{r}
visualized_scheme(TSG_Tier1(), width = 76)
```

```{r}
visualized_scheme(HRR14(), width = 76)
```

## 3.3 `ctdna_variant_filter()`

The single entry point for variant filtering. Two mandatory flags decide what happens:

| `apply` | `explain` | effect |
|:-:|:-:|:--|
| `TRUE`  | `TRUE`  | `prep$variants` filtered; annotated snapshot at `prep$filter_explanation$<sig>` |
| `TRUE`  | `FALSE` | `prep$variants` filtered; no snapshot |
| `FALSE` | `TRUE`  | `prep$variants` unchanged; snapshot only |
| `FALSE` | `FALSE` | error (nothing to do) |

Live `prep$variants` never carries the `filtering_criteria` column; that column lives only on the annotated snapshot.

```{r}
# Apply and explain in one call
prep2 <- ctdna_variant_filter(prep, "TSG_Tier1",
                              apply = TRUE, explain = TRUE, verbose = FALSE)
nrow(prep2$variants)                                             # filtered
nrow(prep2$filter_explanation$TSG_Tier1)                         # all rows, annotated
table(prep2$filter_explanation$TSG_Tier1$filtering_criteria, useNA = "always")
```

Combining schemes — pass a vector of names or a list of schemes. Row passes if any in-scope scheme passes; the criteria column names every in-scope scheme that fired and, for schemes with named branches, which branch fired.

```{r}
prep3 <- ctdna_variant_filter(prep,
  filter_scheme = c("TSG_Tier1","TSG_Tier2","HRR14"),
  apply = TRUE, explain = TRUE, verbose = FALSE)

names(prep3$filter_explanation)   # snapshot key = "TSG_Tier1_TSG_Tier2_HRR14"
head(sort(table(prep3$filter_explanation[[1]]$filtering_criteria,
                useNA = "ifany"), decreasing = TRUE), 6)
# Typical labels:
#   PASS::TSG_Tier1(somatic_nonsyn), TSG_Tier2(somatic_LoF)
#   PASS::TSG_Tier1(deleterious_LGR)
#   FAIL::TSG_Tier1, TSG_Tier2
#   NA   (row in no scheme's scope)
```

## 3.4 `ctdna_create_scheme()`

Thin builder for user schemes. Covers the flat AND/OR case with positive predicates. For anything more complex (negation, nesting) use `create_filtering_scheme()` with the full grammar.

```{r}
my <- ctdna_create_scheme(
  name     = "mini_tsg",
  genes    = c("TP53","RB1"),
  criteria = c("somatic","truncating"),
  combine  = "all",
  overwrite = TRUE)

prep4 <- ctdna_variant_filter(prep, "mini_tsg",
                              apply = TRUE, explain = FALSE, verbose = FALSE)
nrow(prep4$variants)
```

## 3.5 `create_filtering_scheme()` — the full grammar

For schemes that need negation or nested logic. This is the low-level builder; `ctdna_create_scheme()` and every built-in scheme are wrappers around it.

```{r}
sch <- create_filtering_scheme(
  rule(Gene = c("TP53","RB1","PTEN")),
  not(rule_clinvar_benign()),
  not(rule_ch()),
  anyOf(
    allOf(rule_germline(), rule_clinvar_path()),
    allOf(rule_somatic(),
          anyOf(rule_truncating(),
                allOf(rule_missense(), rule_rare_001())))),
  name = "custom_tsg",
  description = "TSG: germline P/LP OR somatic LoF OR somatic rare missense.",
  overwrite = TRUE)
class(sch)
```

## 3.6 Scheme catalog helpers

```{r}
head(ctdna_filter_schemes_list(), 10)
ctdna_filter_scheme_show("HRR14", print = FALSE)$name
```

Persistence — save a scheme to disk and reload it in a later session:

```{r, eval = FALSE}
ctdna_filter_scheme_save(sch, file = "custom_tsg.rds")
sch2 <- ctdna_filter_scheme_load("custom_tsg.rds")
```

# 4. Genomic landscape plots

## 4.1 `ctdna_oncoprint()`

Takes a `prep`, a named list of gene sets, and returns an oncoprint (ComplexHeatmap engine if installed, ggplot fallback otherwise). Baseline visit by default.

```{r, fig.height = 4.5}
# Add an Indication column so wrap works
prep$adrs <- data.frame(
  Patient_ID = prep$clinical$Patient_ID,
  indication = rep(c("NSCLC","HNSCC","SCLC","mCRPC"),
                   length.out = nrow(prep$clinical)))

op <- ctdna_oncoprint(prep,
        gene_sets     = list(TSG   = c("TP53","RB1","PTEN"),
                             HRR14 = ctdna_gene_set("HRR14")),
        filter_scheme = "TSG_Tier1",
        engine        = "ggplot",
        title         = "Baseline landscape by TSG / HRR14")
op$plot
```

## 4.2 `ctdna_alteration_grid()`

Alteration prevalence grid. Rows are gene sets, columns are the RECIST-derived response strata.

Gene-set entries can be bare vectors (default: any-of) or wrapped with `alt_any()`, `alt_all()`, or `alt_any_n(n)` to change how multi-gene sets combine.

```{r, fig.height = 3.5}
gr <- ctdna_alteration_grid(prep,
        gene_sets = list(
          any_TSG    = c("TP53","RB1","PTEN"),
          all_TSG    = alt_all(c("TP53","RB1","PTEN")),
          two_of_TSG = alt_any_n(c("TP53","RB1","PTEN"), n = 2)),
        scheme    = "three",
        title     = "TSG alteration prevalence by response")
gr$plot
```

# 5. Pipeline

`ctdna_pipeline()` runs every deliverable step whose input modalities are present. Result access is uniform: `res$<key>$plot`. A dry run reports what would execute.

```{r}
dry <- ctdna_pipeline(prep, dry_run = TRUE, verbose = FALSE)
str(dry$skipped, max.level = 1)
```

```{r}
res <- ctdna_pipeline(prep,
         which     = c("oncoprint","alteration_grid"),
         scheme    = "three",
         gene_sets = list(TSG   = c("TP53","RB1","PTEN"),
                          HRR14 = ctdna_gene_set("HRR14")),
         engine    = "ggplot",
         verbose   = FALSE)
names(res)
```

# 6. Reference: rule predicates

The rule library exports around 60 predicate helpers. The most common ones:

- Somatic status: `rule_somatic()`, `rule_germline()`, `rule_ch()`
- ClinVar: `rule_clinvar_path()`, `rule_clinvar_benign()`, `rule_clinvar_vus()`
- Molecular consequence: `rule_truncating()`, `rule_missense()`, `rule_synonymous()`, `rule_nonsense()`, `rule_frameshift()`, `rule_inframe_indel()`, `rule_splice_acceptor()`, `rule_splice_donor()`
- Variant type: `rule_is_SNV()`, `rule_is_Indel()`, `rule_is_CNV()`, `rule_is_Fusion()`, `rule_is_LGR()`
- CNV subtypes: `rule_cnv_homozyg_del()`, `rule_cnv_loh_del()`, `rule_cnv_focal_amp()`, `rule_cnv_aneuploid_amp()`
- LGR subtypes: `rule_lgr_deletion()`, `rule_lgr_tandem_duplication()`, `rule_lgr_inversion()`, `rule_is_LGR_deleterious()`
- Population frequency: `rule_rare_001()`, `rule_rare_0001()`, `rule_common_001()`
- Gene sets: `rule_genes_HRR14()`, `rule_genes_TSG()`, `rule_genes_RTK()`, `rule_genes_MMR()`
- Indication: `rule_ind_NSCLC()`, `rule_ind_BRCA()`, `rule_ind_SCLC()`, `rule_ind_HNSCC()`, `rule_ind_CRC()`, `rule_ind_mCRPC()`, `rule_ind_GBM()`

See `?ctdna_rules_library` for the complete list.

# 7. Package configuration

```{r}
ctdna_opts("subject")           # column name for patient ID
ctdna_opts("gene")              # column name for gene symbol
```

Options are set with `ctdna_opts(key = value, ...)` and can be persisted with `ctdna_opts_save()` / `ctdna_opts_load()`.

# Session info

```{r}
sessionInfo()
```
